---
title: "Fermat distances and their supporting graphs on surfaces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fermat distances and their supporting graphs on surfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment='#>',fig.width=7,fig.height=5,warning=FALSE,message=FALSE)
available <- all(vapply(c('grip','igraph','plotly'),requireNamespace,logical(1),quietly=TRUE)) &&
  'distance.matrix' %in% names(formals(grip::metric.mds))
available <- available && file.exists(system.file('examples','fermat-surfaces.R',package='dgraphs')) &&
  'return.graph' %in% names(formals(dgraphs::fermat.distances)) &&
  'distance.matrix' %in% names(formals(grip::metric.mds))
knitr::opts_chunk$set(eval=available)
```

```{r dependencies,echo=FALSE,eval=TRUE,results='asis'}
if(!available) cat('This executable example requires development versions of dgraphs and grip with the direct distance and supporting-graph APIs, plus igraph and plotly. Install these dependencies to render the calculated examples.\n')
```

Fermat distances allow paths through every observation. A symmetric nearest-neighbor
graph is an optional restriction, not part of the complete-graph estimator. This
example asks how that restriction changes distances on an area-uniform paraboloid,
an area-uniform saddle, and an arc-length-uniform helix. The companion grip vignette
focuses on embedding the same distances into three dimensions.

## Generate the observations

The quadratic surfaces are $(u,v,u^2+v^2)$ (index zero) and
$(u,v,u^2-v^2)$ (index one), over the unit disk. Index counts negative eigenvalues
of the quadratic form. We reuse the existing dgraphs area sampler:

```{r sampler}
library(dgraphs)
geometry <- synthetic.quadform(2, 3, forms=list(diag(c(1,-1))))
sampling <- synthetic.sampling.quadform.lab('disk',extent=1,mode='area')
sample <- sample.synthetic.geometry(geometry,sampling,n=150,seed=4101)
X <- sample$predictors
```

The sampler weights parameter-space proposals by
$\sqrt{1+\|2Au\|^2}$, the area element for $(u,u^TAu)$, where $u$ is a
two-dimensional parameter and $A$ is its quadratic-form matrix. This is uniform
on surface area, not on the parameter disk. The helix is
$(\cos t,\sin t,0.2t)$ for $0\leq t\leq4\pi$; uniform $t$ is uniform arc length
because its speed is the constant $\sqrt{1+0.2^2}$.

## Distances and the path graph

For observation indices $i,j$ and exponent $p\geq1$, we compute
\[
D_p(i,j)=\min_{\pi:i\leadsto j}\sum_{(a,b)\in\pi}\|X_a-X_b\|^p.
\]
The minimum ranges over paths through the observed points, including direct edges.
The distance is unrooted, so its units are length to power $p$.

```{r distances}
result <- fermat.distances(points=X,p=2,backend='implicit',rooted=FALSE,
                          return.graph=TRUE)
D <- result$distances
graph <- result$graph
head(graph.edges(graph))
result$metadata
```

The returned graph is the union of one selected shortest-path tree per source.
Its edges carry the powered ambient step lengths. With every source requested,
its shortest-path distances preserve the full matrix. With only landmark sources,
only landmark-to-sample preservation is guaranteed. Ties may admit other valid
unions, and the union need not be sparse: at $p=1$, direct Euclidean edges usually
remain necessary.

A Fermat symmetric kNN graph has different edges: they summarize Fermat paths to
the $k$ nearest neighbors, then take the union of directed neighbor selections.

```{r sknn}
g <- create.fermat.sknn.graph(X,k=5,p=2,rooted=FALSE)
```

## Compare with surface geodesics

The shared example uses powers 1--4 and k values 3--10. Numerical references on
80 reproducibly selected pairs use the existing quadratic-surface geodesic solver:
Clairaut's method for the paraboloid and geodesic shooting for the saddle. Twelve
pairs also use another continuous solver with tighter tolerances. These are
numerical references, not certified globally shortest paths. Helix references
are exact arc lengths $\sqrt{1.04}|t_i-t_j|$.

We compare estimates $d_r$ with reference distances $g_r$ on valid pairs $r$:
\[
a=\frac{\sum_r d_rg_r}{\sum_r d_r^2},\qquad
E=\sqrt{\frac{\sum_r(ad_r-g_r)^2}{\sum_r g_r^2}}.
\]
The fitted global scale $a$ makes powers comparable; $E=0$ means proportional
agreement. It does not establish equality in physical units. Spearman correlation
measures rank agreement. For graph restriction error we compare graph distances
with $D_p$ using the same formula with $a=1$, on all pairs in the largest component.
Disconnected graphs are not repaired, and retained sample counts are reported.
If a retained component contains no reference pair, its geodesic error is undefined
and omitted from the corresponding curve.

```{r shared, include=FALSE}
source(system.file('examples','fermat-surfaces.R',package='dgraphs'))
cache.root <- Sys.getenv('FERMAT_SURFACE_RESULTS')
results <- lapply(c('paraboloid','saddle','helix'),function(shape) {
  path <- file.path(cache.root,paste0(shape,'_4101.rds'))
  if(nzchar(cache.root) && file.exists(path)) return(readRDS(path))
  s <- fermat.surface.sample(shape,150L,4101L)
  fermat.surface.experiment(s,fermat.surface.reference(s))
})
metrics <- do.call(rbind,lapply(results,`[[`,'records'))
```

```{r distance-table, echo=FALSE}
knitr::kable(metrics[metrics$k==0,c('shape','p','n','distance_scale',
 'distance_nrmse','distance_spearman')],digits=3,
 col.names=c('Example','p','Samples','Scale','Scaled error','Rank correlation'))
```

Table 1. Complete-graph Fermat distances compared with the surface or curve
reference. Each row uses the same reference pairs within its example; these are
finite-sample comparisons, not estimates of an asymptotic convergence rate.

```{r graph-error, echo=FALSE}
old <- par(mfrow=c(1,3),mar=c(4,4,2,1))
for(shape in unique(metrics$shape)) {
 z<-metrics[metrics$shape==shape & metrics$k>0,]
 plot(range(z$k),c(0,1.3*max(c(0,z$restriction_nrmse),na.rm=TRUE)),type='n',xlab='Neighbors k',
      ylab='Graph restriction error',main=shape)
 for(p in 1:4){q<-z[z$p==p,];lines(q$k,q$restriction_nrmse,col=p,pch=p,type='b')}
 legend('topright',legend=paste('p =',1:4),col=1:4,lty=1,pch=1:4,bty='n',cex=.7)
}
par(old)
```

**Figure 1.** Distance change caused by retaining only symmetric Fermat kNN edges.
Zero means that the restricted graph preserves the dense Fermat distances on the
retained component. The corresponding component counts and retained sample counts
are available below; a small error on a small component would not demonstrate
whole-sample preservation.

```{r coverage, echo=FALSE}
z<-metrics[metrics$components>1,]
knitr::kable(aggregate(cbind(retained,components)~shape+p,z,range))
```

The table summarizes ranges among disconnected k values within each example and
power. Complete coverage is 150 samples; restricting a disconnected graph can
leave far fewer samples even when its within-component distance error is small.

## Inspect supporting edges in three dimensions

The following view uses dense-target SGD metric MDS from grip. The gray patch is
the underlying surface; colored points are the embedding after one global
similarity alignment for display, and thin blue segments are supporting path
edges. Gray points identify the original observations. Similarity alignment uses
translation, rotation/reflection and uniform scale, not a deformation of the fit.

```{r overlay, echo=FALSE}
fermat.surface.overlay(results[[2]],p=2,k=0)
```

**Figure 2.** Saddle embedding and complete-graph Fermat supporting paths. Drag to
rotate. The legend states how many edges are shown; a deterministic display subset
is used above 1,800 edges. The saved graph contains every selected edge.

Intrinsic distances measure travel along a surface. Euclidean distances in a
three-dimensional embedding measure straight chords, so geometric recovery of the
original surface is not implied by a good distance fit. In particular, exact
helix arc lengths embed on a straight line. See the grip companion for the full
embedding comparison and additional overlays.
