---
title: "Density-sensitive paths with Fermat distances"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density-sensitive paths with Fermat distances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(dgraphs)
```

## The estimator and its graph restriction

The sample Fermat distance between observed points minimizes the sum of
powered Euclidean step lengths over **all sequences of observed points**:

\[
D_p(i,j)=\min_{\pi:i\leadsto j}\sum_{(a,b)\in\pi}\|x_a-x_b\|^p.
\]

`fermat.distances(points = X, p = 2)` computes this complete-graph reference.
It does not first remove long edges. For `p > 1`, several short steps can cost
less than a single long step. This favors densely sampled routes without
explicitly fitting a density estimate. For `p = 1`, the complete-graph answer
is Euclidean distance, not distance around a circle.

The default result is the unrooted sum. `rooted = TRUE` takes its `1/p` power:
this changes reported distances but not minimizing paths. Neither option
applies sample-size normalization or estimates a limiting constant.

```{r small-example}
X <- matrix(c(0, 1, 2), ncol = 1)
fermat.distances(points = X, p = 2)
```

The distance from 0 to 2 is 2: the two unit steps cost less than the direct
step, which costs 4.

Two explicit restrictions are also available:

```{r restrictions}
fermat.distances(points = X, graph.type = "sknn", k = 1, p = 2)
graph <- create.sknn.graph(X, k = 1, neighbor.method = "exact",
                          prune.edges = FALSE, connect.components = FALSE)
fermat.distances(graph, p = 2)
```

Coordinate kNN construction uses union symmetrization: either endpoint may
select an edge. Supplied graphs use their stored lengths as base lengths before
powering. With the same retained weights, restricting edges cannot decrease
the distance. Missing routes return `Inf`; no connectivity repair is implicit.
Use `vertices = c(i, j)` for a selected submatrix; paths still use other vertices.

## Noisy circles with different sampling densities

We generate 100 observations around a unit circle with radial Gaussian noise
of standard deviation 0.04. One example has uniform angular sampling; the
other places half the observations near angle zero. Both retain a uniform
angular component. These are fixed illustrations, not a simulation study or a
proof of geometric recovery. Radial noise changes the sampled geometry from
an exact circle; circle arc length is therefore not a universal truth target.

```{r circle-data}
set.seed(4101)
n <- 100
angles <- list(
  Uniform = sort(runif(n, 0, 2*pi)),
  Concentrated = sort(c(runif(n/2, 0, 2*pi),
                        rnorm(n/2, 0, 0.4) %% (2*pi)))
)
radial.noise <- rnorm(n, 0, 0.04)
circles <- lapply(angles, function(theta) {
  cbind((1 + radial.noise)*cos(theta), (1 + radial.noise)*sin(theta))
})
powers <- c(1, 2, 3)
reference <- lapply(circles, function(X) {
  lapply(powers, function(p) fermat.distances(points = X, p = p))
})
```

To visualize the minimizing routes, the next chunk uses igraph on the same
complete graph and powered Euclidean weights. It checks each plotted path's
cost against `fermat.distances()`. Black points mark the selected endpoints;
red segments show a shortest path. Tied minimizers may have different shapes.
Notice how the preferred route changes with the power and sampling pattern;
a dense route need not be the shortest circular arc.

```{r paths, fig.height=5.5, fig.cap="Complete-graph shortest paths at powers 1, 2 and 3. Rows show uniform and concentrated angular sampling; axes are the observed Cartesian coordinates.", fig.alt="Six circle panels compare complete-graph paths for three powers and two angular sampling distributions. Endpoints are black, observations gray, and minimizing paths red."}
old <- par(mfrow = c(2, 3), mar = c(3, 3, 2, 0.5))
for (s in seq_along(circles)) {
  X <- circles[[s]]
  ends <- c(which.min(rowSums(sweep(X, 2, c(1,0), "-")^2)),
            which.min(rowSums(sweep(X, 2, c(-1,0), "-")^2)))
  g <- igraph::make_full_graph(n, directed = FALSE)
  e <- igraph::as_edgelist(g, names = FALSE)
  eu <- as.matrix(dist(X))
  for (j in seq_along(powers)) {
    p <- powers[j]
    weights <- eu[e]^p
    path <- as.integer(igraph::shortest_paths(g, from = ends[1], to = ends[2],
                                             weights = weights)$vpath[[1]])
    cost <- sum(eu[cbind(head(path, -1), tail(path, -1))]^p)
    stopifnot(isTRUE(all.equal(cost, reference[[s]][[j]][ends[1],ends[2]])))
    plot(X, asp = 1, pch = 16, cex = 0.5, col = "gray60",
         xlab = "x", ylab = "y", main = paste(names(circles)[s], "p =", p))
    lines(X[path, , drop = FALSE], col = "firebrick", lwd = 2)
    points(X[ends, ], pch = 16, cex = 1.1)
  }
}
par(old)
```

## Does a sparse graph give the same distances?

For each circle and power, compare symmetric kNN restrictions at `k = 4, 8, 16`
against **all unordered pairs** of the complete reference. The table reports
unreachable pairs separately. Relative error is measured only for finite
restricted distances; it does not hide disconnections. Numerical agreement
uses tolerance `1e-10 * max(1, reference distance)`.

```{r comparison}
rows <- list()
for (s in seq_along(circles)) for (j in seq_along(powers)) {
  full <- reference[[s]][[j]]
  use <- upper.tri(full)
  for (k in c(4, 8, 16)) {
    restricted <- fermat.distances(points = circles[[s]], p = powers[j],
                                   graph.type = "sknn", k = k)
    finite <- use & is.finite(restricted)
    stopifnot(all(restricted[use] >= full[use] - 1e-10))
    agree <- is.finite(restricted[use]) &
      abs(restricted[use] - full[use]) <= 1e-10*pmax(1, full[use])
    rows[[length(rows)+1L]] <- data.frame(
      Sampling = names(circles)[s], p = powers[j], k = k,
      Unreachable = sum(use & !is.finite(restricted)),
      Agreement_percent = 100*mean(agree),
      Max_relative_error = if (any(finite))
        max(abs(restricted[finite] - full[finite])/full[finite]) else NA_real_
    )
  }
}
knitr::kable(do.call(rbind, rows), digits = 4)
```

Agreement in this table establishes equivalence only for these finite samples,
powers and neighborhood sizes. In particular, the `p = 1` rows demonstrate
that circle-following kNN distances can differ from the Euclidean reference.
Increasing `k` to `n - 1` always restores the complete graph, at quadratic
storage cost. Merely observing no change between two smaller `k` values does
not certify equivalence to the reference.

## What the theory does and does not establish

Groisman, Jonckheere and Sapienza define the unrooted complete-graph estimator
and establish a normalized continuum limit under their sampling assumptions.
They also define directed nearest-neighbor restrictions and prove a
high-probability equality with the unrestricted estimator when `k` exceeds
a bound of the form `c1 * log(n / epsilon) + c2`, under the stated hypotheses
(Definitions 2.1 and 2.11, Proposition 2.12). This does not select a certified
finite-sample `k` for these noisy examples.

Union-symmetric kNN contains the directed edges, so equality for the directed
restriction implies equality for the union restriction by inclusion of paths.
Arbitrary pruning, mutual-neighbor intersections and other supplied graphs
need separate justification. A rooted output is a transformation of the
unrooted estimator and must be interpreted with its corresponding scaling.

This example tests graph restriction directly rather than assuming it is
harmless. Further assessment should use multiple sample sizes and seeds, and
should distinguish recovering ordinary geometry from deliberately changing
distances through sampling density. Neither a visually appealing path nor a
connected graph establishes the continuum theorem's assumptions.

Complete-graph construction and all-pairs output both require quadratic
storage. Both coordinate modes currently evaluate all Euclidean pairs; kNN
restriction reduces the stored graph, not that neighbor-search work. Start
with small reference examples. Positive lengths that overflow or underflow
to zero after powering are rejected; rescale the coordinates if necessary.

## References

Groisman, P., Jonckheere, M. and Sapienza, F. (2022).
*Nonhomogeneous Euclidean first-passage percolation and distance learning.*
Bernoulli 28(1), 255--276. [Published paper](https://doi.org/10.3150/21-BEJ1341)
and [preprint with definitions and propositions](https://arxiv.org/html/1810.09398v2).

For ordinary graph distances and graph construction, see
[Constructing and Diagnosing Data-Derived Graphs](data-derived-graph-workflow.html).
