---
title: "Finding your way around dgraphs"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Finding your way around dgraphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
body { max-width: 1100px; }
td, th { vertical-align: top; }
td code { white-space: normal; overflow-wrap: anywhere; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

`dgraphs` turns numerical observations into graphs and provides tools to inspect
their connectivity, paths and geometry. Start with the question you want to
answer. Use the [task map](#choose-a-starting-point), follow a small workflow,
then consult the [complete catalog](#function-catalog).

## Choose a starting point

| Your question | Entry point | Choice that matters |
|---|---|---|
| Which nearby observations should be linked? | `create.mknn.graph()` or `create.sknn.graph()` | Mutual neighbors require agreement; symmetric neighbors accept either direction. |
| How should support adapt to sampling density? | `create.rknn.graph()` or `create.cknn.graph()` | Fixed radius, local radius rules, or a geometric mean of local scales. |
| Should overlapping neighborhoods define edges? | [`create.single.iknn.graph()`](#help-create.single.iknn.graph) | Intersection graphs can connect points that are not direct nearest neighbors. |
| How sensitive is my graph to neighborhood size? | Plural constructors such as `create.sknn.graphs()` | Inspect a parameter sequence and its component counts. |
| Do I need paths between every pair? | `graph.connected.components()`, then constructor repair controls | Bridge edges make a modeling assumption across gaps. |
| How far apart are vertices along this graph? | [`graph.geodesic.distances()`](#help-graph.geodesic.distances) | Edge values must represent additive lengths or costs. |
| Does the graph preserve known distances? | [`summarize.isometry.deviation()`](#help-summarize.isometry.deviation) | Requires a reference distance matrix with matching vertex order. |
| Do I need coordinates for a graph? | `graph.embedding()` or `graph.spectral.embedding()` | A visualization layout and a spectral representation serve different purposes. |
| How can I generate points with known geometry? | [`sample.synthetic.geometry()`](#help-sample.synthetic.geometry) | Pair a geometry with a compatible sampling specification. |

## A first graph

Use a finite numeric matrix with observations in rows. This example samples a
circle, constructs a symmetric nearest-neighbor graph, requests connectivity
repair, and measures distances on the final graph.

```{r first-graph}
library(dgraphs)
points <- sample.synthetic.geometry(
  synthetic.circle(),
  synthetic.sampling.grid.interval(0, 2 * pi, endpoints = "exclude.lower"),
  n = 24, seed = 17
)
X <- points$predictors
graph <- create.sknn.graph(X, k = 2, connect.components = TRUE)
print(graph)
length(unique(graph.connected.components(graph)))
D <- graph.geodesic.distances(graph)
round(D[1:4, 1:4], 3)
ig <- as_igraph(graph)
c(vertices = igraph::vcount(ig), edges = igraph::ecount(ig))
```

Distances here sum straight chords between observed coordinates. They
approximate circular arc distances through a finite graph; they are not exact
continuous arc lengths. For a longer comparison of graph families, repair and
diagnostics, read [Constructing and Diagnosing Data-Derived Graphs](data-derived-graph-workflow.html).
For sampling and embedding details, read [Synthetic geometry and point sampling](synthetic-geometry.html).

## Constructing a standard graph

Use `create.graph(type, ...)` when you want a familiar topology or a small
example. The required `type` selects the graph; arguments following `n` must
be named. Every result uses the same graph accessors as a data-derived graph.

| Type | Arguments | Edge lengths |
|---|---|---|
| `"empty"` | `n` (zero allowed) | Absent |
| `"complete"` | `n` | Absent |
| `"chain"` | `n` or `x`; optional `span = 1`, `y` | One, or absolute coordinate differences |
| `"cycle"` | `n >= 3` | Absent |
| `"complete_bipartite"` | `sizes = c(n1, n2)` | Absent |
| `"star"` | `arms`: positive edge counts along each arm | Absent |
| `"random"` | `n`, `mean.degree`; optional `connected = TRUE`, `seed` | Independent uniform lengths on [0.5, 1.5] |
| `"circle"` | `n >= 3`; optional `sampling`, `edge.length`, `seed` | Shorter unit-circle arcs (default) or chords |

```{r standard-graphs}
standard <- list(
  empty = create.graph("empty", n = 5),
  complete = create.graph("complete", n = 5),
  chain = create.graph("chain", n = 5, span = 2, labels = 101:105),
  cycle = create.graph("cycle", n = 5),
  complete_bipartite = create.graph("complete_bipartite", sizes = c(2, 3)),
  star = create.graph("star", arms = c(1, 2, 1)),
  random = create.graph("random", n = 5, mean.degree = 2, seed = 23),
  circle = create.graph("circle", n = 5, sampling = "uniform",
                        edge.length = "chord")
)
knitr::kable(data.frame(
  type = names(standard),
  vertices = vapply(standard, graph.order, integer(1)),
  edges = vapply(standard, function(g) nrow(graph.edges(g)), integer(1)),
  stored.lengths = vapply(standard, function(g) !is.null(graph.lengths(g)), logical(1))
), row.names = FALSE)
```

For chains, `span` counts positions on **each side**, so an interior vertex
has at most `2 * span` neighbors. This differs from `k` in nearest-neighbor
workflows. Optional `x` sorts vertices by coordinate, with ties resolved by
input position; `y` and `labels` follow the same ordering. The sorted-to-input
mapping is `metadata$order`. An ordinary star has all arm counts equal to one;
larger counts subdivide its arms.

A cycle describes topology. A circle also records unit-circle coordinates in
`metadata$coordinates` and angles in `metadata$angles`. Circle sampling is
`"random"` by default; `"uniform"` spaces angles equally. Choose
`edge.length = "chord"` for Euclidean edge lengths or `"arc"` for shorter
circular arcs. A shorter arc can cross an unsampled gap; graph shortest paths
still follow the edges actually present. Arc lengths are measured in radians
on the unit circle, not squared distances or conductances.

Random graphs have `floor(n * mean.degree / 2)` edges. A connected graph needs
at least `n - 1`, and an impossible request errors. With `connected = TRUE`,
a growing random tree precedes additional edges; the result is not a uniform
draw from all connected graphs. With `connected = FALSE`, edges are sampled
uniformly without replacement from all vertex pairs. Candidate storage is
quadratic in the number of vertices, so this generator is intended for small
examples. For both random graphs and randomly sampled circles, an explicit
`seed` restores the caller's random state after generation; `seed = NULL`
uses and advances the current stream. Uniform circle sampling uses no random
numbers. Unknown or irrelevant arguments error instead of being ignored.

## Conventions that affect your choice

### A common graph object

Every graph constructor returns a `dgraph`, including basic fixtures. Use
`graph.adjacency(graph)` for its neighbor lists, `graph.lengths(graph)` for
aligned lengths, and `graph.edges(graph)` for one row per undirected edge.
`graph.order(graph)` includes isolated vertices. A bare `adj.list` has one
integer vector per vertex, containing **1-based** neighbor indices;
`length.list` gives the corresponding finite, nonnegative lengths.

`dgraph(adj.list, length.list)` validates reciprocal edges and values,
rejects duplicates and self-loops, and sorts neighbors and values together.
Omit `length.list` for topology alone. A stored zero-length edge remains an
edge. `as_igraph(graph)` preserves isolates, lengths as the `length` attribute,
and other named edge attributes. For igraph distance calculations, explicitly
use `weights = igraph::E(ig)$length`. Raw-list matrix conversions retain
`weight.list` for quantities that need not represent distances.

Path-search results contain their graph in `$graph` and paths in
`$shortest.paths`; a parameter sequence contains separate graph objects.
Subgraphs use compact indices and record `metadata$original.vertices`.
`create.graph(type, ..., labels = ...)` assigns vertex labels independently
of adjacency indices. For example, `labels = 101:110` names ten vertices
without changing their internal indices `1:10`.

### Neighborhood sizes and edge quantities

All sequence constructors take an explicit, strictly increasing `k.values`
vector with `1 <= k < n`. In coordinate and graph-geodesic intersection
workflows, **k counts other vertices**. Each cover explicitly includes its
source plus those neighbors. Equal distances are resolved by vertex index,
including coincident observations. Coordinate ANN search uses an exact
fallback when its boundary candidates are tied. The specialized
`knn.metric = "linf.simplex"` search currently evaluates all candidates to
honor the same tie rule, with quadratic distance-evaluation cost.

```{r intersection-convention}
X.line <- matrix(c(0, 1, 3, 6), ncol = 1)
initial <- create.single.iknn.graph(X.line, k = 1, prune.method = "none",
                                  connect.components = FALSE, verbose = FALSE)
rebuilt <- create.geodesic.iknn.graph(initial, k = 1)
rebuilt$metadata$neighborhood
```

Graph-geodesic construction errors if any component has fewer than `k + 1`
vertices. Choose `small.component = "truncate"` to use all reachable other
vertices instead; `metadata$neighborhood$effective.k` records the count for
each vertex, including zero for isolates. Iterated construction applies the
same requested k and component policy at every rebuild.

| Construction | Edge-length contract |
|---|---|
| Mutual and symmetric kNN | Euclidean distance between endpoints. |
| Fixed/adaptive radius and continuous kNN | Euclidean distance; radius rules govern which edges exist. |
| Coordinate intersection kNN | Minimum summed metric distance through a common cover member, in working coordinates after any PCA. |
| Geodesic intersection kNN | Previous-graph shortest-path distance between the new endpoints. |
| Nerve of arbitrary sets | No lengths; cardinality of overlap is the `overlap` edge attribute. |

An overlap score or conductance is not a travel distance. Read named
attributes with `graph.edge.attribute(graph, "overlap")`, or the corresponding
column of `graph.edges(graph)`. A length-based operation rejects a graph with
no lengths; `graph.geodesic.distances(graph, distance = "hop")` explicitly
requests unweighted distance. `graph.spectrum(graph)` uses the unweighted
Laplacian. `graph.embedding(graph)` is unweighted by default; choose an
`edge.attribute` and `transform = "identity"` or `"reciprocal"` explicitly.
Layout values must be strictly positive, so zero lengths require a separate,
explicit choice of layout quantity.

### Connectivity, stages and cost

Mutual, symmetric and radius constructors can leave components separate.
Request `connect.components = TRUE` when bridges are appropriate. Exact
component-MST repair joins components through shortest Euclidean bridges;
the ANN candidate version can fall back to exact search. A global minimum
spanning tree is another repair option. `join.graphs()` identifies one vertex
from each of two graphs and preserves aligned lengths and attributes.

`graph.stages(graph)` lists the stored stages. Every graph has `final`;
constructors may additionally retain `raw`, `pruned`, `raw.repaired`,
`pruned.repaired` and `repaired.pruned`. Pass the same `stage` to any accessor,
`as_igraph()` or `graph.geodesic.distances()`. A missing stage errors, and
stage selection never prunes, repairs or reconstructs a graph. The geodesic
intersection constructor stores only `final` because it performs no pruning
or repair. Symmetric kNN's `graph.detail = "minimal"` requires both disabled.

Disconnected pairs have infinite distances. All-pairs distances require a
quadratic-size matrix; request selected vertices when possible. Geometric
pruning and exact component bridging can be expensive even with ANN neighbor
search. Native construction is serial; thread controls do not promise native
parallel execution.

### Distances and stored routes

Use `graph.geodesic.distances()` for a matrix of distances. Supply exactly one
of an existing `graph` or a coordinate matrix `points`. With coordinates,
`graph.type = "sknn"` requires an explicit `k` and builds a symmetric kNN graph
without pruning or connectivity repair. `graph.type = "mst"` builds a Euclidean
minimum spanning tree and does not accept `k`. Setting `k = 1` never changes
the construction method. Coincident observations remain connected by
zero-length edges when selected by the construction.

```{r distance-inputs}
line.points <- matrix(c(0, 0, 1, 10, 11), ncol = 1)
D.tree <- graph.geodesic.distances(points = line.points, graph.type = "mst")
D.neighbors <- graph.geodesic.distances(points = line.points,
                                       graph.type = "sknn", k = 1)
knitr::kable(data.frame(
  pair = c("Coincident observations", "First to last observation"),
  tree.distance = c(D.tree[1, 2], D.tree[1, 5]),
  symmetric.neighbor.distance = c(D.neighbors[1, 2], D.neighbors[1, 5])
), row.names = FALSE, col.names = c("Pair", "Tree distance", "Symmetric kNN distance"))
```

Both calculations are exact shortest-path distances on their chosen graphs.
Neither construction guarantees exact continuous manifold geodesics. The
symmetric-neighbor example has disconnected components, represented by `Inf`.
`vertices` selects matrix rows and columns in the requested order; paths may
still pass through any vertex in the graph. `distance = "hop"` counts edges;
`distance = "length"` requires stored lengths. To reuse a graph or choose
additional construction options, construct it first and pass `graph = g`.

Use `create.path.graph()` when the routes themselves matter. Its `h.values`
argument specifies strictly increasing positive integer **maximum hop counts**;
a hop traverses one edge. It always returns a named collection, including
when only one limit is requested. Select a member before retrieving a route.

```{r constrained-paths}
weighted.graph <- dgraph(
  list(c(2L, 3L), c(1L, 3L, 4L), c(1L, 2L), 2L),
  list(c(5, 1), c(5, 1, 1), c(1, 1), 1))
paths <- create.path.graph(weighted.graph, h.values = c(2, 3))
knitr::kable(compare.paths(paths, from = 1, to = 4)[,
  c("h", "path_exists", "path_length", "n_hops")], row.names = FALSE,
  col.names = c("Hop limit", "Route exists", "Route length", "Hops used"))
get.shortest.path(paths[["h_2"]], from = 1, to = 4)
get.shortest.path(paths[["h_3"]], from = 4, to = 1)
```

With two hops, the best route has length 6; allowing three hops lowers it to 3.
The returned route's edges, stored length and hop count agree. Each result
stores a graph of reachable pairs in `$graph`, with a `hops` edge attribute.
`get.shortest.path()` returns a vertex sequence, length and hop count, or
`NULL` when the pair is unreachable within the selected limit. Both directions
are supported; a vertex's route to itself has length and hop count zero.
For equal-length routes, fewer hops are preferred. These all-pairs route
collections can require substantial memory; use modest hop limits when possible.

For a path expressed as coordinates, `path.length()` sums Euclidean segment
lengths and `subdivide.path()` includes both endpoints at equally spaced
distances along the polyline. Points on opposite sides of a corner need not
have equal straight-line distances from each other.

```{r coordinate-path}
polyline <- rbind(c(0, 0), c(1, 0), c(1, 2))
path.length(polyline) # 3
subdivide.path(polyline, n.subdivision.pts = 5)
path.dist(1:3, polyline) # normalized cumulative distances: 0, 1/3, 1
```

The five subdivision rows correspond to arc distances 0, 0.75, 1.5, 2.25 and 3.
Repeated consecutive points contribute zero length. For a singleton or a path
whose points all coincide, length and normalized cumulative distances are zero,
and subdivision repeats that location. Empty paths are rejected; subdivision
requires at least two output points.

### Positive spectra and spectral coordinates

`graph.spectrum()` uses the final graph's unweighted combinatorial Laplacian,
regardless of stored lengths. Its `nev` counts **positive eigenpairs**, ordered
from smallest upward. Every connected component, including an isolated vertex,
contributes one omitted zero mode. The spectrum records this count as `nullity`
and preserves component IDs in vertex order. `nev = NULL` requests all positive
modes; use a small explicit count for large graphs. Requests larger than the
number of positive modes error rather than silently changing the count.

```{r positive-spectrum}
chain <- create.graph("chain", n = 8)
spectrum <- graph.spectrum(chain, nev = 3)
round(spectrum$evalues, 6)
spectrum$nullity
coordinates <- graph.spectral.embedding(spectrum, dim = 2)
dim(coordinates)
```

The three eigenvalues are approximately 0.152241, 0.585786 and 1.234633; the
embedding uses the first two. `use.R = TRUE` has the same output contract.
Both backends honor `return.Laplacian = TRUE` and `return.dense`; sparse output
uses the optional Matrix package. The native backend uses a sparse solver for
partial spectra and a dense solver for the full spectrum. The R backend uses
a full dense solve, so both an R solve and a full native spectrum require
quadratic memory. A documented absolute tolerance separates computed zero
modes from positive modes; an unresolved separation errors explicitly.

Pass the spectrum object to `graph.spectral.embedding()`. Its default
`scale = "none"` uses orthonormal eigenvectors; `scale = "inverse.sqrt"`
scales each column by the inverse square root of its eigenvalue. With every
positive mode on a connected graph, squared distances in the latter embedding
equal effective resistance. With fewer modes it is a truncated representation.
Neither coordinate choice generally reproduces graph shortest-path distances.
Signs and bases in repeated eigenspaces can differ between solvers. Include
the whole tied eigenspace when comparing distances between embeddings.

Disconnected graphs have spectra, but no common embedding is assigned: select
a component with `create.subgraph(graph, which(spectrum$components == id))`,
compute its spectrum and embed it separately. Distances between independently
embedded components have no defined meaning. Empty and singleton graphs have
no positive-dimensional spectral embedding.

### Plotting a graph and reusing its layout

`graph.embedding()` computes coordinates without drawing. The standard
`plot()` method for a `dgraph` computes a two-dimensional layout only when
coordinates are absent and invisibly returns the coordinates actually drawn.
Use `vertex.values` for numeric values mapped through a palette, or
`vertex.colors` for explicit colors; supplying both errors. Constant numeric
values use the palette midpoint. Every undirected edge is drawn once.

```{r graph-layout, fig.width=6, fig.height=4, fig.alt="Graph layout with vertex colors defined by the example; the same coordinates are reused for comparison."}
cycle <- create.graph("cycle", n = 8)
set.seed(4101)
xy <- plot(cycle, vertex.values = seq_len(8), main = "Numeric vertex values")
```

```{r graph-layout-reuse, fig.width=6, fig.height=4, fig.alt="Graph layout with vertex colors defined by the example; the same coordinates are reused for comparison."}
plot(cycle, coordinates = xy, vertex.values = rep(7, 8),
     main = "Same layout, constant values")
```

The same `stage` controls both layout computation and drawn edges. For
three-dimensional coordinates, choose an explicit two-column projection;
the plotting method never silently discards a coordinate. Layout generation
itself remains available in both two and three dimensions.

```{r graph-projection, fig.width=6, fig.height=4, fig.alt="Graph layout with vertex colors defined by the example; the same coordinates are reused for comparison."}
set.seed(4102)
xyz <- graph.embedding(cycle, dim = 3)
colnames(xyz) <- c("x", "y", "z")
plot(cycle, coordinates = xyz, projection = c(1, 3), vertex.colors = "navy",
     main = "Explicit x/z projection")
```

### Reuse an expensive computation

Keep a constructed graph for repeated `graph.geodesic.distances()` calls;
this reuses its edges, not a memoized distance matrix. Save `D` yourself when
reusing the same distances. Coordinate input constructs a fresh graph for
each distance call. Geodesic intersection and iterated intersection
workflows deliberately build on existing graph metrics.

`create.sknn.graphs()` reuses one ANN search across its `k` values. A single
symmetric constructor also accepts ordered, 1-based, non-self `knn.index`
rows with `neighbor.method = "ann"`; these must correspond to the same input
points. Intersection constructors expose `knn.cache.path` and
`knn.cache.mode`: `"read"` requires a valid cache, `"write"` replaces it, and
`"readwrite"` creates a missing cache but rejects an existing invalid one.
Format version 3 caches store only non-self neighbors and check point values,
row order, metric, working-coordinate preprocessing and maximum k. Old cache
formats are rejected. The adaptive-radius sequence has a batched ANN backend. These are distinct
contracts; there is no universal public neighbor-cache object shared by all
constructors. Internal cache and native solver helpers are not installed
public functions to call directly.

## Function catalog

The catalog covers all 112 explicit public exports exactly once;
44 registered S3 methods are described separately through object workflows.
Internal helpers and development applications are not public entry points.
Links open the exact help included with this build, offline.

Each row is one explicit export. Open exact arguments and examples with
`help("create.sknn.graph", package = "dgraphs")`, replacing the name as needed.
The [package reference manual](https://CRAN.R-project.org/package=dgraphs/dgraphs.pdf)
describes the released version; newly added functions require the corresponding
release or development installation.

### Graph construction and parameter sequences

| Function | Purpose |
|---|---|
| [`create.mknn.graph()`](#help-create.mknn.graph) | Build a mutual Euclidean nearest-neighbor graph, with optional pruning and repair. |
| [`create.mknn.graphs()`](#help-create.mknn.graphs) | Build mutual-neighbor graphs across a range of neighborhood sizes. |
| [`create.sknn.graph()`](#help-create.sknn.graph) | Build a symmetric Euclidean nearest-neighbor graph, with optional pruning and repair. |
| [`create.sknn.graphs()`](#help-create.sknn.graphs) | Build a symmetric-neighbor sequence using a shared ANN search. |
| [`create.rknn.graph()`](#help-create.rknn.graph) | Build a fixed-radius or adaptive-radius Euclidean graph. |
| [`create.rknn.graphs()`](#help-create.rknn.graphs) | Vary the local neighborhood scale in adaptive-radius graphs. |
| [`create.cknn.graph()`](#help-create.cknn.graph) | Build a continuous-kNN graph using geometric-mean local scales. |
| [`create.single.iknn.graph()`](#help-create.single.iknn.graph) | Build one intersection-neighborhood graph with optional pruning. |
| [`create.iknn.graphs()`](#help-create.iknn.graphs) | Build an intersection-neighborhood sequence and its pruning branches. |
| [`create.geodesic.iknn.graph()`](#help-create.geodesic.iknn.graph) | Rebuild intersection neighborhoods using an existing graph's distances. |
| [`create.iterated.iknn.graphs()`](#help-create.iterated.iknn.graphs) | Repeat intersection construction in successive graph metrics. |
| [`build.iknn.graphs.and.selectk()`](#help-build.iknn.graphs.and.selectk) | Select an intersection neighborhood size using structural or label-mixing stability. |
| [`create.cmst.graph()`](#help-create.cmst.graph) | Complete a Euclidean minimum spanning tree using an edge-length quantile. |
| [`create.threshold.distance.graph()`](#help-create.threshold.distance.graph) | Connect points satisfying a distance threshold. |
| [`nerve.graph()`](#help-nerve.graph) | Construct a graph from overlaps between sets in a cover. |

### Basic graphs and representation

| Function | Purpose |
|---|---|
| [`create.graph()`](#help-create.graph) | Construct standard graph types, labeled chains, random graphs and sampled circles. |
| [`dgraph()`](#help-dgraph) | Validate aligned neighbor, length and attribute lists as an undirected graph. |
| [`graph.adjacency()`](#help-graph.adjacency) | Read the neighbor list for a stored stage. |
| [`graph.lengths()`](#help-graph.lengths) | Read aligned lengths, or NULL when absent. |
| [`graph.edges()`](#help-graph.edges) | Read one ordered row per undirected edge, with lengths and attributes. |
| [`graph.order()`](#help-graph.order) | Count vertices, including isolates. |
| [`graph.stages()`](#help-graph.stages) | List stages actually stored in a graph. |
| [`graph.edge.attribute()`](#help-graph.edge.attribute) | Read an explicitly named edge attribute. |
| [`as_igraph()`](#help-as_igraph) | Convert a dgraph stage to an undirected igraph object. |
| [`convert.adjacency.list.to.adjacency.matrix()`](#help-convert.adjacency.list.to.adjacency.matrix) | Convert bare neighbor lists to an adjacency matrix. |
| [`convert.weighted.adjacency.matrix.to.adjacency.list()`](#help-convert.weighted.adjacency.matrix.to.adjacency.list) | Convert a weighted adjacency matrix to aligned lists. |
| [`convert.adjacency.to.edge.matrix()`](#help-convert.adjacency.to.edge.matrix) | Extract an edge matrix and optional aligned weights. |
| [`count.edges()`](#help-count.edges) | Count undirected edges once from reciprocal neighbor lists. |
| [`compare.adj.lists()`](#help-compare.adj.lists) | Compare neighbor sets, optionally reporting differences. |
| [`edge.diff()`](#help-edge.diff) | Find edges present in one graph but absent from another. |
| [`identical.vertex.set.weighted.graph.similarity()`](#help-identical.vertex.set.weighted.graph.similarity) | Compare weighted graphs on the same vertex set. |

### Connectivity and graph editing

| Function | Purpose |
|---|---|
| [`graph.connected.components()`](#help-graph.connected.components) | Assign a component identifier to every vertex. |
| [`convert.to.undirected()`](#help-convert.to.undirected) | Add reciprocal links to a directed adjacency list. |
| [`rm.self.loops()`](#help-rm.self.loops) | Remove each vertex from its own neighbor list. |
| [`join.graphs()`](#help-join.graphs) | Identify one vertex from each of two graphs. |
| [`create.subgraph()`](#help-create.subgraph) | Extract the graph induced by selected vertices. |
| [`create.grid.graph()`](#help-create.grid.graph) | Refine graph edges toward approximately uniform spatial spacing. |

### Paths and distances

| Function | Purpose |
|---|---|
| [`graph.geodesic.distances()`](#help-graph.geodesic.distances) | Compute shortest-path distances for a supported graph object and stage. |
| [`compute.graph.distance()`](#help-compute.graph.distance) | Compute a weighted shortest-path distance between two vertices. |
| [`compute.graph.diameter()`](#help-compute.graph.diameter) | Compute a weighted graph diameter. |
| [`get.shortest.path()`](#help-get.shortest.path) | Recover a shortest vertex path. |
| [`geodesic.disk()`](#help-geodesic.disk) | Find a graph neighborhood within a distance radius. |
| [`geodesic.knn()`](#help-geodesic.knn) | Find nearest vertices in a graph metric. |
| [`geodesic.knnx()`](#help-geodesic.knnx) | Find graph-metric nearest neighbors for query vertices. |
| [`create.path.graph()`](#help-create.path.graph) | Construct a collection of path results for one or several hop limits. |
| [`create.plm.graph()`](#help-create.plm.graph) | Build a path-length-matrix graph structure. |
| [`compare.paths()`](#help-compare.paths) | Compare paths obtained under different hop limits. |
| [`minh.limit()`](#help-minh.limit) | Find a minimum hop limit that permits a path. |
| [`path.length()`](#help-path.length) | Measure Euclidean length along a coordinate path. |
| [`path.dist()`](#help-path.dist) | Compute normalized cumulative distance along a vertex path. |
| [`subdivide.path()`](#help-subdivide.path) | Include both endpoints and space points equally along a polyline; repeat the location for zero-length paths. |

### Edge lengths and pruning

| Function | Purpose |
|---|---|
| [`extract.trajectory.edge.lengths()`](#help-extract.trajectory.edge.lengths) | Extract stored edge lengths along a graph vertex trajectory. |
| [`wgraph.prune.long.edges()`](#help-wgraph.prune.long.edges) | Prune edges using length constraints on alternative paths. |

### Diagnostics, endpoints and packing

| Function | Purpose |
|---|---|
| [`compute.graph.summary.pmf()`](#help-compute.graph.summary.pmf) | Represent a degree, weight, component-size or label-neighborhood summary as probability masses. |
| [`graph.summary.divergence()`](#help-graph.summary.divergence) | Compare graph-summary probability masses using a chosen divergence. |
| [`jensen.shannon.divergence()`](#help-jensen.shannon.divergence) | Measure Jensen-Shannon divergence between probability distributions. |
| [`compute.graph.summary.stability()`](#help-compute.graph.summary.stability) | Track summary divergence between consecutive graphs. |
| [`compute.stability.metrics()`](#help-compute.stability.metrics) | Summarize structural stability of an intersection-graph sequence. |
| [`compute.geodesic.stats()`](#help-compute.geodesic.stats) | Summarize geodesic growth around grid vertices. |
| [`compute.vertex.geodesic.stats()`](#help-compute.vertex.geodesic.stats) | Inspect geodesic growth around one grid vertex. |
| [`compute.graph.endpoint.scores()`](#help-compute.graph.endpoint.scores) | Score directional concentration in a 3D embedding using graph neighborhoods. |
| [`detect.graph.endpoints()`](#help-detect.graph.endpoints) | Select and consolidate candidate geometric endpoints. |
| [`geodesic.core.endpoints()`](#help-geodesic.core.endpoints) | Detect endpoints using a graph-geodesic core. |
| [`detect.local.extrema()`](#help-detect.local.extrema) | Find graph-local maxima or minima of vertex values. |
| [`vertices()`](#help-vertices) | Extract vertices through the generic, currently for local-extrema results. |
| [`remove.knn.outliers()`](#help-remove.knn.outliers) | Identify and remove observations with unusual nearest-neighbor distances. |
| [`create.maximal.packing()`](#help-create.maximal.packing) | Select separated vertices at a requested graph scale. |
| [`validate.maximal.packing()`](#help-validate.maximal.packing) | Check separation and maximality of a vertex packing. |
| [`verify.maximal.packing()`](#help-verify.maximal.packing) | Verify a packing result using its graph and scale. |
| [`isometry.scale()`](#help-isometry.scale) | Fit a global scale aligning estimated distances with reference distances. |
| [`isometry.rel.rms.error()`](#help-isometry.rel.rms.error) | Measure target-normalized root-mean-square distance error. |
| [`isometry.rel.abs.error()`](#help-isometry.rel.abs.error) | Summarize quantiles of relative absolute distance errors. |
| [`isometry.distortion.quantiles()`](#help-isometry.distortion.quantiles) | Summarize multiplicative distance distortion. |
| [`isometry.distance.correlations()`](#help-isometry.distance.correlations) | Compare estimated and reference distances by correlation. |
| [`isometry.geodesic.diagnostics()`](#help-isometry.geodesic.diagnostics) | Diagnose shortcuts and distance-scale-dependent geodesic bias. |
| [`summarize.isometry.deviation()`](#help-summarize.isometry.deviation) | Combine scale, error, distortion and correlation diagnostics. |

### Spectral methods and graph display

| Function | Purpose |
|---|---|
| [`graph.spectrum()`](#help-graph.spectrum) | Compute the smallest positive Laplacian eigenpairs, with component nullity. |
| [`graph.spectral.embedding()`](#help-graph.spectral.embedding) | Embed a connected graph from its spectrum, with explicit coordinate scaling. |
| [`graph.embedding()`](#help-graph.embedding) | Compute a graph visualization layout. |

### Synthetic geometry and embedding

| Function | Purpose |
|---|---|
| [`synthetic.quadform()`](#help-synthetic.quadform) | Specify a flat or quadratic geometry with intrinsic and ambient dimensions. |
| [`synthetic.circle()`](#help-synthetic.circle) | Specify a circle or circular arc. |
| [`synthetic.helix()`](#help-synthetic.helix) | Specify a helical curve. |
| [`synthetic.trefoil()`](#help-synthetic.trefoil) | Specify the supported trefoil parameterization. |
| [`synthetic.sphere.cap()`](#help-synthetic.sphere.cap) | Specify a spherical cap over a planar footprint. |
| [`synthetic.torus.patch()`](#help-synthetic.torus.patch) | Specify a torus patch over two angular parameters. |
| [`synthetic.simplex()`](#help-synthetic.simplex) | Specify a compositional simplex with a given number of parts. |
| [`synthetic.stratum.point()`](#help-synthetic.stratum.point) | Specify a zero-dimensional point stratum. |
| [`synthetic.stratum.segment()`](#help-synthetic.stratum.segment) | Specify a one-dimensional line-segment stratum. |
| [`synthetic.stratum.rectangle()`](#help-synthetic.stratum.rectangle) | Specify a two-dimensional rectangular stratum. |
| [`synthetic.stratified()`](#help-synthetic.stratified) | Assemble compatible strata and optional junction metadata; general sampling is not implemented. |
| [`synthetic.point.line.junction()`](#help-synthetic.point.line.junction) | Specify a point and segment junction. |
| [`embed.synthetic.geometry()`](#help-embed.synthetic.geometry) | Deterministically map supported latent coordinates to ambient coordinates. |
| [`embed.quadform.surface()`](#help-embed.quadform.surface) | Lift two-column coordinates using the versioned Geometry Lab quadratic arithmetic. |
| [`quadform.gradient()`](#help-quadform.gradient) | Evaluate gradients of quadratic height functions. |
| [`quadform.metric()`](#help-quadform.metric) | Evaluate the induced metric of a quadratic geometry. |
| [`edge.lengths.synthetic.geometry()`](#help-edge.lengths.synthetic.geometry) | Measure lifted straight latent segments; currently supports quadforms only. |
| [`validate.synthetic.geometry()`](#help-validate.synthetic.geometry) | Validate a geometry specification or latent coordinate matrix. |

### Point sampling

| Function | Purpose |
|---|---|
| [`sample.synthetic.geometry()`](#help-sample.synthetic.geometry) | Realize geometry-only points with an explicit seed or random-state plan. |
| [`synthetic.sampling.uniform.interval()`](#help-synthetic.sampling.uniform.interval) | Specify uniform interval draws and their ordering. |
| [`synthetic.sampling.grid.interval()`](#help-synthetic.sampling.grid.interval) | Specify deterministic interval points and endpoint inclusion. |
| [`synthetic.sampling.uniform.box()`](#help-synthetic.sampling.uniform.box) | Specify uniform box draws with scalar or coordinate-wise bounds. |
| [`synthetic.sampling.uniform.rectangle()`](#help-synthetic.sampling.uniform.rectangle) | Specify two-dimensional rectangle draws. |
| [`synthetic.sampling.uniform.disk()`](#help-synthetic.sampling.uniform.disk) | Specify parameter-area-uniform disk draws. |
| [`synthetic.sampling.truncated.normal()`](#help-synthetic.sampling.truncated.normal) | Specify bounded normal draws on an interval. |
| [`synthetic.sampling.gapped.uniform()`](#help-synthetic.sampling.gapped.uniform) | Specify draws across disjoint intervals with allocation controls. |
| [`synthetic.sampling.clustered()`](#help-synthetic.sampling.clustered) | Specify a fixed number of two-dimensional clusters and observations per cluster. |
| [`synthetic.sampling.dirichlet.zeros()`](#help-synthetic.sampling.dirichlet.zeros) | Specify compositional draws with a fraction on a structural-zero face. |
| [`synthetic.sampling.quadform.lab()`](#help-synthetic.sampling.quadform.lab) | Specify versioned quadratic-surface sampling, including surface-area rejection sampling. |
| [`validate.synthetic.sampling()`](#help-validate.synthetic.sampling) | Validate a sampling specification and its compatibility with a geometry. |

### Interactive graph reconstruction explorer

| Function | Purpose | Input and interpretation |
|---|---|---|
| [`explore.graphs()`](#help-explore.graphs) | Open the Graph Reconstruction Explorer. | A saved benchmark directory, manifest, or registered project name; optional Shiny dependencies. |
| [`read.graph.benchmark()`](#help-read.graph.benchmark) | Read saved benchmark indexes without launching an app. | Existing dggraphui manifests and asset tables; relative paths resolve within the run. |
| [`register.graph.project()`](#help-register.graph.project) | Add a benchmark to the local project picker. | Stores a name and path; scientific assets remain outside the package. |

Opening a project reads saved results without refitting. An explicit action can
compute a missing weighted layout using grip and save it to a separate user cache.
The default display preserves original coordinates with equal x/y/z units.
A centered common-scale view preserves proportions; optional per-axis scaling
is labeled as distorting proportions. Saved metric tables never change with the display.

Try the installed, deterministic twelve-point circle project:

```{r explorer-demo}
demo <- system.file("extdata", "graph-explorer-demo", package = "dgraphs")
benchmark <- read.graph.benchmark(demo)
nrow(benchmark$graph_assets)
```

```{r explorer-launch, eval=FALSE}
explore.graphs(demo)
```

Compare k = 2 with k = 4. The additional chords shorten graph distances
relative to exact shorter unit-circle arcs. The saved GRIP layouts visualize
the graphs; their apparent circularity is not the distance-error metric.

## Object workflows and method help

Call a generic such as `print(x)`, `summary(x)` or `plot(x)`; R selects its
registered S3 method from the object's class. A class need not support all
three operations. For example, print a symmetric graph, summarize a mutual
sequence, or plot an intersection stability result. `vertices(extrema)`
extracts vertices from `detect.local.extrema()` output. Geometry samples have concise print and summary methods; plot their
coordinate matrices explicitly. Path-series summaries compare hop limits in a table.

The method links open version-matched help below. From R, the same topics
are available through `help("print.sknn_graph", package = "dgraphs")`.
Links inside another package's help may require online access; dgraphs topics
and their examples are included here and work without a network connection.

| Generic | Object class and detailed method help |
|---|---|
| `print()` | [`dgraph`](#help-print.dgraph); [`adaptive_radius_graph`](#help-print.adaptive_radius_graph); [`build_iknn_graphs_and_selectk`](#help-print.build_iknn_graphs_and_selectk); [`cknn_graph`](#help-print.cknn_graph); [`geodesic_stats`](#help-print.geodesic_stats); [`knn.outliers`](#help-print.knn.outliers); [`maximal_packing`](#help-print.maximal_packing); [`mknn_graph`](#help-print.mknn_graph); [`mknn_graphs`](#help-print.mknn_graphs); [`mst_completion_graph`](#help-print.mst_completion_graph); [`packing_validation`](#help-print.packing_validation); [`path.graph`](#help-print.path.graph); [`path.graph.plm`](#help-print.path.graph.plm); [`radius_graph`](#help-print.radius_graph); [`rknn_graphs`](#help-print.rknn_graphs); [`sknn_graph`](#help-print.sknn_graph); [`sknn_graphs`](#help-print.sknn_graphs); [`summary.knn.outliers`](#help-print.summary.knn.outliers); [`summary.local_extrema`](#help-print.summary.local_extrema); [`summary.mst_completion_graph`](#help-print.summary.mst_completion_graph); [`synthetic_geometry_sample`](#help-print.synthetic_geometry_sample); [`summary.synthetic_geometry_sample`](#help-print.summary.synthetic_geometry_sample); [`path.graph.series`](#help-print.path.graph.series) |
| `summary()` | [`IkNN`](#help-summary.IkNN); [`geodesic_stats`](#help-summary.geodesic_stats); [`iknn_graphs`](#help-summary.iknn_graphs); [`iterated_iknn_graphs`](#help-summary.iterated_iknn_graphs); [`knn.outliers`](#help-summary.knn.outliers); [`local_extrema`](#help-summary.local_extrema); [`mknn_graphs`](#help-summary.mknn_graphs); [`mst_completion_graph`](#help-summary.mst_completion_graph); [`path.graph`](#help-summary.path.graph); [`rknn_graphs`](#help-summary.rknn_graphs); [`vertex_geodesic_stats`](#help-summary.vertex_geodesic_stats); [`synthetic_geometry_sample`](#help-summary.synthetic_geometry_sample); [`path.graph.series`](#help-summary.path.graph.series) |
| `plot()` | [`dgraph`](#help-plot.dgraph); [`IkNNgraphs`](#help-plot.IkNNgraphs); [`build_iknn_graphs_and_selectk`](#help-plot.build_iknn_graphs_and_selectk); [`cst_graph_mixing_stats`](#help-plot.cst_graph_mixing_stats); [`geodesic_stats`](#help-plot.geodesic_stats); [`iknn_stability_metrics`](#help-plot.iknn_stability_metrics); [`vertex_geodesic_stats`](#help-plot.vertex_geodesic_stats) |
| [`vertices()`](#help-vertices) | [`local_extrema`](#help-vertices.local_extrema) |

## Finding more help

```{r help, eval=FALSE}
help(package = "dgraphs")
help("dgraphs-package", package = "dgraphs")
vignette(package = "dgraphs")
```

The organization draws inspiration from the short function-purpose entries
and related-help navigation in [Hmisc's overview](https://CRAN.R-project.org/package=Hmisc/refman/Hmisc.html#HmiscOverview)
and its [usage examples](https://hbiostat.org/r/hmisc/).
The task groups here follow dgraphs' own API. The source repository's
`make audit-guides` target checks catalog completeness and method-help coverage
against `NAMESPACE` and the maintained help sources.

## Detailed help for this build

These expandable topics are generated from the installed package's help
sources. They are the same documentation opened by `help()` in R.

```{r offline-reference, echo=FALSE, results='asis'}
source(system.file("doc-tools", "reference.R", package = "dgraphs"), local = TRUE)
render_dgraphs_reference()
```

<style>
.offline-help { margin: 1em 0; padding: .7em; border: 1px solid #ccd7dd; }
.offline-help summary { cursor: pointer; font-weight: 600; }
:focus-visible { outline: 3px solid #176a8a; outline-offset: 3px; }
</style>
