---
title: "dgraphs: Constructing, Selecting, and Diagnosing Data-Derived Graphs in R"
date: "2026-08-20"
abstract: >
  Data-derived graphs turn numerical observations into vertices and local
  proximity relations into edges. Small choices in the construction rule can
  alter connectivity, graph geodesics, and downstream spectral analyses.
  The dgraphs package provides mutual and symmetric k-nearest-neighbor graphs,
  shared- and intersection-neighbor graphs, fixed and adaptive radius graphs,
  continuous-kNN graphs, graph-geodesic iterations, and minimum-spanning-tree
  completion in one R interface. Constructors preserve native, pruned, and
  connectivity-repaired lifecycle stages instead of silently replacing the
  original graph. Parameter sequences, structural stability summaries,
  geodesic diagnostics, path utilities, spectral embeddings, and conversion
  to igraph support an inspectable workflow from construction to analysis. We
  describe the object design, demonstrate the workflow on a variable-density
  circular point cloud, verify symmetric-kNN edge parity with FNN and dbscan,
  and compare four construction families in a controlled geodesic experiment.
  The package does not select a universally correct graph: its purpose is to
  make graph assumptions and their consequences explicit and reproducible.
draft: true
author:
  - name: Pawel Gajer
    affiliation: University of Maryland School of Medicine
    address:
      - Center for Advanced Microbiome Research and Innovation
      - Institute for Genome Sciences
      - Department of Microbiology and Immunology
      - Baltimore, Maryland, USA
    email: pgajer@gmail.com
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{float}
  \usepackage{placeins}
packages:
  cran:
    - dgraphs
    - FNN
    - dbscan
    - igraph
    - ggplot2
  bioc: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.7,
  fig.align = "center",
  out.width = "100%",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png",
  dpi = 150
)

library(dgraphs)
library(ggplot2)

stopifnot(packageVersion("dgraphs") >= "0.1.0.9000")

pipeline.raw <- utils::read.csv("data/pipeline-benchmark.csv")
family.raw <- utils::read.csv("data/graph-family-benchmark.csv")

pipeline.summary <- stats::aggregate(
  elapsed_seconds ~ backend + n + dimension,
  pipeline.raw,
  stats::median
)
names(pipeline.summary)[names(pipeline.summary) == "elapsed_seconds"] <-
  "median_seconds"

family.summary <- stats::aggregate(
  cbind(
    edges,
    components_before_repair,
    bridges_added,
    normalized_mae,
    shortcut_fraction,
    distance_correlation
  ) ~ family + n + density + noise,
  family.raw,
  stats::median
)

set.seed(20260820)
n.example <- 80L
theta <- sort(c(
  stats::runif(55L, 0, pi),
  stats::runif(25L, pi, 2 * pi)
))
X <- cbind(x = cos(theta), y = sin(theta)) +
  matrix(stats::rnorm(2L * n.example, sd = 0.015), ncol = 2L)
```

# Introduction

A data-derived graph has one vertex for each observation and edges determined
from distances or neighbor relations. Such graphs are inputs to manifold
learning, spectral methods, clustering, trajectory reconstruction, and local
diagnostics. The construction affects the analysis: a fixed-radius graph can
fragment in sparse regions, a symmetric k-nearest-neighbor (kNN) graph can add
links across density transitions, and a mutual-kNN graph can be conservative
enough to disconnect. Consequently, choosing a constructor and choosing its
scale are parts of the statistical analysis.

Several R packages efficiently find nearest neighbors. \CRANpkg{FNN} exposes
kd-tree, cover-tree, and brute-force searches [@fnnPackage], while
\CRANpkg{dbscan} supplies fast spatial indexing together with density-based
clustering methods [@hahsler2019dbscan]. \CRANpkg{igraph} provides a broad
graph-analysis and visualization system [@igraph2023]. These packages are
complementary to \CRANpkg{dgraphs}: the latter focuses on constructing and
comparing data-derived graph families, retaining construction stages, and
diagnosing the geometry of the resulting graph [@dgraphsPackage].

The implemented families have distinct precedents. Shared-nearest-neighbor
ideas trace to @jarvisPatrick1973, mutual-kNN connectivity has been studied for
clustering and outlier detection [@brito1997], and continuous kNN (CkNN) uses
the geometric mean of local scales to adapt to sampling density
[@berrySauer2019]. Optional component repair uses minimum-spanning-tree edges,
linking the graph construction to the classical relationship between minimum
spanning trees and single-linkage structure [@gowerRoss1969]. This article is
about the software design and workflow; it does not propose a new graph rule.

# Graph rules and object lifecycle

Let $x_1,\ldots,x_n \in \mathbb{R}^p$, let $N_k(i)$ be the set of $k$
nearest non-self neighbors of vertex $i$, and let $\sigma_i$ be the distance
from $x_i$ to its $k$th neighbor. The principal undirected rules are:

\[
\begin{array}{ll}
\text{mutual kNN:} & i\sim j \quad\text{if }i\in N_k(j)\text{ and }j\in N_k(i),\\
\text{symmetric kNN:} & i\sim j \quad\text{if }i\in N_k(j)\text{ or }j\in N_k(i),\\
\text{fixed radius:} & i\sim j \quad\text{if }\lVert x_i-x_j\rVert_2\le r,\\
\text{adaptive radius:} & i\sim j \quad\text{if }\lVert x_i-x_j\rVert_2
  \le \delta f(\sigma_i,\sigma_j).
\end{array}
\]

The adaptive rules use `min`, `max`, or geometric-mean choices for $f$; the
geometric-mean case is CkNN. The package also provides intersection-kNN graphs,
which retain information about shared neighborhoods, and iterated
graph-geodesic constructions that replace the ambient distance by shortest
paths on a preceding graph.

Every current constructor returns aligned `adj_list` and `weight_list` fields.
Weights are Euclidean edge lengths unless a documented function states
otherwise. Constructors that support pruning and repair also retain lifecycle
branches:

| Stage | Meaning |
|:--|:--|
| `raw_*` | Native construction before pruning or repair. |
| `pruned_*` | Optional geometric pruning applied to the native graph. |
| `raw_repaired_*` | Native graph after component repair. |
| `pruned_repaired_*` | Prune-first graph after component repair. |
| `repaired_pruned_*` | Repair-first graph after pruning. |
| `adj_list`, `weight_list` | Final graph selected by the constructor call. |

This redundancy is deliberate. A minimum-spanning-tree bridge makes a shortest
path finite but does not make that bridge part of the native neighborhood
model. Keeping both stages permits downstream code to distinguish a support
edge from a repair edge.

# An inspectable construction workflow

We use a noisy circular point cloud with unequal sample counts on the two
semicircles. The known circular ordering supplies a reference geodesic, while
the density imbalance stresses globally scaled constructions.

```{r candidate-graphs}
graphs <- list(
  mutual = create.mknn.graph(
    X, k = 6, connect.components = TRUE
  ),
  symmetric = create.sknn.graph(
    X, k = 6, neighbor.method = "ann", connect.components = TRUE
  ),
  continuous = create.cknn.graph(
    X, k.scale = 6, delta = 1, connect.components = TRUE
  ),
  adaptive.max = create.rknn.graph(
    X, type = "adaptive.radius", k.scale = 6,
    radius.rule = "max", connect.components = TRUE
  )
)
```

```{r candidate-summary, echo=FALSE}
candidate.summary <- do.call(rbind, lapply(names(graphs), function(name) {
  graph <- graphs[[name]]
  data.frame(
    graph = name,
    edges = graph$n_edges,
    raw_components = graph$n_components_before,
    final_components = graph$n_components_after,
    bridges = graph$n_mst_edges_added
  )
}))
names(candidate.summary) <- c(
  "Graph", "Edges", "Raw components", "Final components", "Bridges"
)
knitr::kable(candidate.summary, row.names = FALSE)
```

The edge count is descriptive rather than an objective to maximize. The
component and bridge columns reveal how much of the final connected graph came
from repair.

```{r workflow-figure, echo=FALSE, fig.pos="H", fig.cap="Continuous-kNN graph on a variable-density circular point cloud. Gray segments are final graph edges; blue points are observations. The constructor retains native and repaired stages separately.", fig.alt="A circular point cloud with denser sampling on the upper semicircle. Gray edges connect nearby blue points around the circle."}
selected <- graphs$continuous
edge.matrix <- convert.adjacency.to.edge.matrix(selected$adj_list)$edge.matrix

graphics::plot(
  X,
  asp = 1,
  pch = 19,
  col = "#1F5A94",
  xlab = "Coordinate 1",
  ylab = "Coordinate 2"
)
graphics::segments(
  X[edge.matrix[, 1L], 1L], X[edge.matrix[, 1L], 2L],
  X[edge.matrix[, 2L], 1L], X[edge.matrix[, 2L], 2L],
  col = grDevices::adjustcolor("grey30", alpha.f = 0.42)
)
graphics::points(X, pch = 19, col = "#1F5A94")
```

Plural constructors make parameter sweeps explicit and return a statistics
table rather than requiring users to reconstruct diagnostics from each object.

```{r parameter-sequence, results="hide"}
radius.sequence <- create.rknn.graphs(
  X,
  k.values = 3:7,
  radius.search = "ann",
  connect.components = TRUE
)
radius.statistics <- radius.sequence$k_statistics[, c(
  "k", "n_edges_before_pruning", "n_components_before",
  "n_mst_edges_added", "n_components_after"
)]
```

```{r parameter-table, echo=FALSE}
colnames(radius.statistics) <- c(
  "k", "Native edges", "Native components", "Bridges", "Final components"
)
knitr::kable(radius.statistics, row.names = FALSE)
```

For intersection-kNN sequences, `compute.stability.metrics()` compares
consecutive graphs using edit distance and Jensen--Shannon divergence between
graph-summary distributions. `build.iknn.graphs.and.selectk()` adds explicit
selection rules based on structural stability, label mixing, or both. These
functions return the full curves, chosen indices, and the evidence used for
selection.

# Geodesics, summaries, and interoperability

The final continuous-kNN graph can be converted directly to an `igraph`
object. Conversion preserves isolated vertices, vertex names when available,
edge lengths in the `weight` attribute, and graph-specific attributes such as
intersection size when present.

```{r conversion}
selected.igraph <- as_igraph(selected)
c(
  vertices = igraph::vcount(selected.igraph),
  edges = igraph::ecount(selected.igraph)
)

degree.summary <- compute.graph.summary.pmf(
  selected,
  summary = "degree_distribution"
)
degree.summary$pmf
```

`graph.geodesic.distances()` computes weighted shortest-path distances. For
the circular example, the shorter angular arc is a known reference. The
isometry diagnostics quantify scale-adjusted stress, distance correlation,
shortcut frequency, and bias over distance ranges.

```{r geodesic-diagnostics, results="hide"}
graph.distance <- graph.geodesic.distances(selected)
angle.difference <- abs(outer(theta, theta, "-"))
reference.distance <- pmin(angle.difference, 2 * pi - angle.difference)

geodesic.diagnostics <- round(isometry.geodesic.diagnostics(
  graph.distance,
  reference.distance
), 3)
```

```{r geodesic-table, echo=FALSE}
knitr::kable(
  data.frame(
    Metric = names(geodesic.diagnostics),
    Value = as.numeric(geodesic.diagnostics)
  ),
  row.names = FALSE
)
```

The same representation supports path extraction, endpoint scoring, local
extrema, graph spectra, and spectral embeddings. Their interpretation still
depends on connectivity, repair edges, and geodesic distortion.

# Comparative validation

## Symmetric-kNN edge parity and timing scope

We compared complete symmetric-kNN graph pipelines using `k = 10` on Gaussian
data with 500, 1,500, or 3,000 observations and 2, 10, or 50 dimensions. The
`dgraphs` pipeline used its bundled ANN kd-tree and returned a lifecycle graph
object. The \CRANpkg{FNN} and \CRANpkg{dbscan} pipelines performed kd-tree
queries and then used the same R helper to union directed neighbor pairs and
compute Euclidean edge weights. Each scenario was warmed up and timed three
times. All `r nrow(pipeline.raw)` retained runs reproduced exactly the same
undirected edge set; ties were absent in these continuous random data.

```{r pipeline-timing, echo=FALSE, fig.pos="H", fig.cap="End-to-end symmetric-kNN graph construction times. Points are medians of three runs. The dgraphs pipeline also assembles graph lifecycle diagnostics; the FNN and dbscan pipelines use a common R edge-materialization helper. Times were measured on the R-devel system recorded with the benchmark data.", fig.alt="Faceted line plots of construction time by sample size for dgraphs, FNN, and dbscan in 2, 10, and 50 dimensions. Time increases with sample size and dimension."}
ggplot(
  pipeline.summary,
  aes(x = n, y = median_seconds, color = backend)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~ dimension, labeller = label_both) +
  scale_y_log10() +
  scale_color_manual(values = c(
    dgraphs = "#1F5A94", FNN = "#C44E52", dbscan = "#55A868"
  )) +
  labs(
    x = "Observations",
    y = "Median elapsed seconds (log scale)",
    color = "Pipeline"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

The timing comparison is intentionally scoped to the returned graph, not only
nearest-neighbor lookup. It should not be read as a general benchmark of ANN
libraries. The edge-parity result shows that users can move
between a neighbor-search workflow and the richer `dgraphs` graph object
without changing the symmetric-kNN rule on these tie-free data.

## Construction families under density imbalance

A second experiment used noisy samples from a unit circle, where the true
shorter-arc distance is known. We varied sample size (100 or 200), angular
density (uniform or uneven), and coordinate noise (0.005 or 0.05), with three
replicates per setting. Mutual-kNN, symmetric-kNN, CkNN, and fixed-radius graphs
used a nominal scale of six neighbors. The fixed radius was the median sixth-
neighbor distance. Exact component-MST repair was enabled for every family.

The experiment evaluates sensitivity rather than establishing a general
ordering. Symmetric-kNN produced the fewest disconnected native graphs in
these settings, while the other families more often required repair. Which
graph had the smallest geodesic error depended on density, noise, and sample
size. The raw replicate-level results, generation script, edge-parity checks,
and session metadata accompany the article so that the comparison can be
audited or extended.

```{r family-fidelity, echo=FALSE, fig.pos="!t", fig.cap="Median normalized absolute geodesic error across graph families. Lower values indicate closer agreement with the known circular arc distances. Each point summarizes three replicates; panels separate density and coordinate-noise settings. Connectivity repair was enabled for all graphs.", fig.alt="Four faceted line plots comparing normalized geodesic error for mutual kNN, symmetric kNN, continuous kNN, and fixed-radius graphs at 100 and 200 points under uniform and uneven density and low and high noise."}
ggplot(
  family.summary,
  aes(x = n, y = normalized_mae, color = family)
) +
  geom_line(linewidth = 0.65) +
  geom_point(size = 1.8) +
  facet_grid(noise ~ density, labeller = label_both) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Observations",
    y = "Median normalized absolute error",
    color = "Graph family"
  ) +
  theme_bw(base_size = 9.5) +
  theme(legend.position = "bottom")
```

\FloatBarrier

# Scope and limitations

`dgraphs` assumes finite numerical observations and currently centers its fast
native constructors on Euclidean distance. ANN-backed searches are exact at
the supported tolerance; deterministic ties can still be backend-dependent and
should be tested with `neighbor.method = "exact"` when tie semantics matter.
Some pruning methods are experimental and are labeled as such in the help
pages. Connectivity repair guarantees a connected final graph when possible,
but it does not validate that the added bridges represent scientific
continuity. Likewise, a stable parameter curve does not prove that a graph
recovers a latent manifold.

The package is therefore most useful when graph construction is treated as a
modeling decision. Users should compare lifecycle stages, inspect repair
edges, examine parameter sequences, and evaluate downstream quantities against
domain-specific references or sensitivity analyses.

# Summary and availability

\CRANpkg{dgraphs} integrates graph construction, lifecycle-aware repair,
parameter selection, diagnostics, and `igraph` interoperability in one R
package. The software is available from CRAN and developed at
<https://github.com/pgajer/dgraphs>. The end-to-end package vignette can be
opened with:

```{r vignette-command, eval=FALSE}
vignette("data-derived-graph-workflow", package = "dgraphs")
```

The article source, benchmark scripts, raw CSV results, and citation
verification record are maintained with the package repository.
