---
title: "dgraphs: Constructing and Assessing Data-Derived Graphs in R"
date: "2026-08-24"
abstract: >
  A data-derived graph represents observations by vertices and weighted edges,
  raising the question of how faithfully its shortest-path geometry represents
  the data geometry that an analysis is intended to preserve. The dgraphs
  package constructs mutual and symmetric nearest-neighbor graphs, shared- and
  intersection-neighbor graphs, fixed and adaptive radius graphs,
  continuous-nearest-neighbor graphs, graph-geodesic iterations, and
  minimum-spanning-tree completions. Constructors retain native, pruned, and
  connectivity-repaired stages so that added support edges remain
  distinguishable from the original neighborhood model. The package provides
  target-normalized graph-geodesic relative RMSE, distortion, correlation,
  and distance-band diagnostics, together with parameter sequences, path and
  spectral utilities, and conversion to igraph. We describe the software
  design, demonstrate a construction-and-assessment workflow on a controlled
  circular example, verify edge parity with FNN and dbscan, and compare four
  graph families across sample size, sampling density, and noise. The aim is
  not to prescribe a universally correct graph, but to expose construction and
  repair choices and quantify their geometric consequences.
author:
  - name: Pawel Gajer
    affiliation: University of Maryland School of Medicine
    address:
      - Center for Advanced Microbiome Research and Innovation
      - Institute for Genome Sciences
      - Department of Microbiology and Immunology
      - Baltimore, Maryland, USA
    email: pgajer@gmail.com
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{float}
  \usepackage{placeins}
packages:
  cran:
    - dgraphs
    - FNN
    - dbscan
    - igraph
    - ggplot2
    - knitr
    - rmarkdown
    - rjtools
  bioc:
    - bluster
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.7,
  fig.align = "center",
  out.width = "100%",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png",
  dpi = 150
)

library(dgraphs)
library(ggplot2)

stopifnot(packageVersion("dgraphs") >= "0.2.0")

accessible.kable <- function(data, caption, ...) {
  alt <- knitr::opts_current$get("tab.alt")
  if (!is.character(alt) || length(alt) != 1L || !nzchar(trimws(alt))) {
    stop("Accessible tables require a nonempty tab.alt chunk option.")
  }
  if (knitr::is_latex_output()) {
    return(knitr::kable(data, format = "latex", caption = caption, ...))
  }
  escaped.alt <- gsub("&", "&amp;", alt, fixed = TRUE)
  escaped.alt <- gsub("\"", "&quot;", escaped.alt, fixed = TRUE)
  escaped.alt <- gsub("<", "&lt;", escaped.alt, fixed = TRUE)
  escaped.alt <- gsub(">", "&gt;", escaped.alt, fixed = TRUE)
  knitr::kable(
    data,
    format = "html",
    caption = caption,
    table.attr = paste0('aria-label="', escaped.alt, '"'),
    ...
  )
}

pipeline.raw <- utils::read.csv("data/pipeline-benchmark.csv")
family.raw <- utils::read.csv("data/graph-family-benchmark.csv")

pipeline.summary <- stats::aggregate(
  elapsed_seconds ~ backend + n + dimension,
  pipeline.raw,
  stats::median
)
names(pipeline.summary)[names(pipeline.summary) == "elapsed_seconds"] <-
  "median_seconds"

family.summary <- stats::aggregate(
  cbind(
    native_edges,
    final_edges,
    native_components,
    final_components,
    bridges_added,
    final_rel_rms_error,
    final_shortcut_fraction,
    final_distance_correlation
  ) ~ family + n + density + noise,
  family.raw,
  stats::median
)

scenario.id <- interaction(
  family.summary$n,
  family.summary$density,
  family.summary$noise,
  drop = TRUE
)
family.summary$is.best <- ave(
  family.summary$final_rel_rms_error,
  scenario.id,
  FUN = function(value) value == min(value)
)

family.lifecycle <- do.call(rbind, lapply(
  split(family.raw, family.raw$family),
  function(data) {
    data.frame(
      Family = data$family[[1L]],
      Connected = sum(data$native_components == 1L),
      Components = stats::median(data$native_components),
      Bridges = stats::median(data$bridges_added),
      `Final RelRMSE` = stats::median(data$final_rel_rms_error),
      check.names = FALSE
    )
  }
))
winner.counts <- stats::aggregate(
  is.best ~ family,
  family.summary,
  sum
)
family.lifecycle[["Lowest-loss scenarios"]] <- winner.counts$is.best[
  match(family.lifecycle$Family, winner.counts$family)
]

set.seed(20260820)
n.example <- 80L
theta <- sort(c(
  stats::runif(55L, 0, pi),
  stats::runif(25L, pi, 2 * pi)
))
X <- cbind(x = cos(theta), y = sin(theta)) +
  matrix(stats::rnorm(2L * n.example, sd = 0.015), ncol = 2L)
```

# Introduction

Many analyses first represent observations by a graph and then represent that
graph in low-dimensional coordinates. Fidelity has two corresponding parts:
whether graph geodesics preserve the data dissimilarities of interest, and
whether an embedding preserves the graph's intrinsic geometry. The
\CRANpkg{dgraphs} addresses the first question; the companion \CRANpkg{grip}
package addresses the second [@gripPackage]. Let
$X=\{x_1,\ldots,x_n\}$ denote the observed data, let $G=(V,E,\ell)$ be the
weighted graph constructed on those observations, and let
$Z=\{z_1,\ldots,z_n\}\subset\mathbb{R}^m$, with $m\geq 2$, be a coordinate
configuration for the vertices of $G$. For each vertex pair $i<j$, let
$d_{ij}^X$ be the data dissimilarity of interest and let $d_{ij}^G$ be the
shortest-path distance in $G$. Choose an input-graph geodesic $\gamma_{ij}$
that realizes $d_{ij}^G$, and let $d_{ij}^{G,Z}$ be the length of that selected
geodesic after its vertices are placed at $Z$. We write
$D_X=(d_{ij}^X)$, $D_G=(d_{ij}^G)$, and
$D_{G,Z}=(d_{ij}^{G,Z})$. Depending on the application, the entries of $D_X$
may be observed Euclidean distances, known intrinsic distances in a controlled
example, or domain-specific dissimilarities. These objects define two fidelity
stages:

$$
X
\xrightarrow[
  D_X\ \text{versus}\ D_G
]{
  \substack{\texttt{dgraphs}\\
  \text{construction and assessment}}
}
G
\xrightarrow[
  D_G\ \text{versus}\ D_{G,Z}
]{
  \substack{\texttt{grip}\\
  \text{layout, scoring, and refinement}}
}
Z\subset\mathbb{R}^m,
$$

\noindent where this article concerns $D_X$ versus $D_G$ in the $X\to G$
stage; it does not assess the second transformation.

With the profiled scale
$\alpha=\sum_{i<j}d_{ij}^Gd_{ij}^X/\sum_{i<j}(d_{ij}^G)^2$, which rescales
the graph distances toward the fixed data-distance target, the primary
fidelity loss used here is the target-normalized graph-geodesic relative RMSE

$$
\operatorname{RelRMSE}_{X\to G}
=
\sqrt{
  \frac{\sum_{i<j}(\alpha d_{ij}^G-d_{ij}^X)^2}
       {\sum_{i<j}(d_{ij}^X)^2}
}.
$$

This is the quantity returned by `isometry.rel.rms.error()` and the
`rel_rms_error` field of `summarize.isometry.deviation()`. It is normalized by
the target distances and is not Kruskal's Stress-1. Other package diagnostics
describe correlation, multiplicative distortion, shortcuts, and bias across
distance ranges.

A data-derived graph has one vertex for each observation and edges determined
from distances or neighbor relations. Such graphs are inputs to manifold
learning, spectral methods, clustering, trajectory reconstruction, and local
diagnostics. The construction affects the analysis: a fixed-radius graph can
fragment in sparse regions, a symmetric k-nearest-neighbor (kNN) graph can add
links across density transitions, and a mutual-kNN graph can be conservative
enough to disconnect. Consequently, choosing a constructor and choosing its
scale are parts of the statistical analysis.

Several R packages efficiently find nearest neighbors. \CRANpkg{FNN} exposes
kd-tree, cover-tree, and brute-force searches [@fnnPackage], while
\CRANpkg{dbscan} supplies fast spatial indexing together with density-based
clustering methods [@hahsler2019dbscan]. \CRANpkg{igraph} provides a broad
graph-analysis and visualization system [@igraphPackage]. These packages are
complementary to \CRANpkg{dgraphs}: the latter focuses on constructing and
comparing data-derived graph families, retaining construction stages, and
diagnosing the geometry of the resulting graph [@dgraphsPackage].

The implemented families have distinct precedents. Shared-nearest-neighbor
ideas trace to @jarvisPatrick1973, mutual-kNN connectivity has been studied for
clustering and outlier detection [@brito1997], and continuous kNN (CkNN) uses
the geometric mean of local scales to adapt to sampling density
[@berrySauer2019]. Optional component repair uses minimum-spanning-tree edges,
linking the graph construction to the classical relationship between minimum
spanning trees and single-linkage structure [@gowerRoss1969]. This article is
about the software design and workflow; it does not propose a new graph rule.

# Positioning and design choices

\CRANpkg{dgraphs} is intended for analysts and method developers whose downstream
result depends on the graph construction itself. Its role begins after a data
matrix has been assembled and ends with an explicit graph, its construction
history, and diagnostics that can be passed to another method. Users who need
only nearest-neighbor indices can use \CRANpkg{FNN}; users focused on
density-based clustering can work directly with the kNN, fixed-radius, and
shared-nearest-neighbor objects in \CRANpkg{dbscan} [@hahsler2019dbscan]. The
Bioconductor package \BIOpkg{bluster} constructs kNN and shared-neighbor
`igraph` objects for community detection, allows the neighbor-search backend to
be selected through BiocNeighbors, and supports threaded construction
[@blusterPackage]. These are closer alternatives than a general graph library.
\CRANpkg{dgraphs} differs by treating comparisons among graph rules, construction
stages, repair order, and geodesic consequences as part of the returned
analysis rather than as preprocessing for one clustering algorithm.

The public interface follows a `create.<family>.graph()` convention, with
plural constructors for parameter sequences. Constructor results are S3 lists
rather than a new external graph class. Aligned adjacency and weight lists keep
isolated vertices and intermediate branches directly inspectable, serialize
with base R, and make it possible to compare native and repaired graphs without
reconstructing them. The tradeoff is that retaining several lifecycle branches
uses more memory than returning one final edge set. `as_igraph()` provides the
boundary to the broader graph ecosystem instead of reproducing analysis
algorithms already available in \CRANpkg{igraph}.

R performs argument validation, result assembly, and lifecycle bookkeeping;
nearest-neighbor search and the main construction kernels use C or C++ through
Rcpp [@eddelbuettelFrancois2011]. The runtime dependencies are limited to FNN,
igraph, parallel, and Rcpp, with RcppEigen used at compilation
[@batesEddelbuettel2013]. The primary constructors are currently
single-process, and the benchmarks in this article use one process. A few
stand-alone utilities expose worker counts, but graph construction itself does
not. Thus, `bluster` is a better fit when threaded kNN or shared-neighbor graph
construction is the main requirement, while \CRANpkg{dgraphs} favors deterministic
cross-family diagnostics and lifecycle retention.

For $n$ observations in $p$ dimensions and $m$ retained edges, the input matrix
requires $O(np)$ storage and one adjacency/weight branch requires $O(n+m)$.
Parameter sequences and lifecycle retention increase memory in proportion to
the total number of stored branches and edges. ANN-backed Euclidean searches
avoid an explicit all-pairs distance matrix; exact reference paths and exact
component bridging can require quadratic pair comparisons. Diagnostics that
return all pairwise graph distances necessarily allocate an $n$-by-$n$ result.
For larger data, users can select ANN-backed constructors, request minimal graph
detail when supported, and avoid retaining parameter sequences or all-pairs
diagnostics that are not needed.

The public API is organized around four recurring tasks. Table 1 lists a
representative subset rather than every exported method.

```{r api-map-table, echo=FALSE, tab.alt="Four-row map of the dgraphs public API. The rows group representative functions for graph construction and parameter sequences, lifecycle pruning and repair, data-versus-graph assessment, and conversion, path, and spectral analysis."}
api.break <- if (knitr::is_latex_output()) " \\\\ " else "<br>"
api.code <- function(value) {
  if (knitr::is_latex_output()) {
    value <- gsub("_", "\\_", value, fixed = TRUE)
    paste0("\\texttt{", value, "}")
  } else {
    paste0("<code>", value, "</code>")
  }
}
api.wrap <- function(...) {
  value <- paste(vapply(c(...), api.code, character(1L)), collapse = api.break)
  if (knitr::is_latex_output()) paste0("\\shortstack[l]{", value, "}") else value
}
api.map <- data.frame(
  Task = c(
    "Construction and parameter sequences",
    "Lifecycle, pruning, and repair",
    "Data-to-graph assessment",
    "Conversion, paths, and spectra"
  ),
  `Representative functions` = c(
    api.wrap("create.sknn.graph()", "create.rknn.graphs()"),
    api.wrap("wgraph.prune.long.edges()", "create.cmst.graph()"),
    api.wrap("graph.geodesic.distances()", "summarize.isometry.deviation()"),
    api.wrap("as_igraph()", "get.shortest.path()", "graph.spectrum()")
  ),
  check.names = FALSE
)
accessible.kable(
  api.map,
  caption = paste(
    "Representative public functions grouped by task. Constructors return",
    "lifecycle-aware graph objects; assessment functions compare data",
    "dissimilarities with graph geodesics; conversion and analysis utilities",
    "connect those objects to downstream graph workflows."
  ),
  row.names = FALSE,
  escape = FALSE,
  position = "H"
)
```

# Graph rules and object lifecycle

Let $x_1,\ldots,x_n \in \mathbb{R}^p$, let $N_k(i)$ be the set of $k$
nearest non-self neighbors of vertex $i$, and let $\sigma_i$ be the distance
from $x_i$ to its $k$th neighbor. The principal undirected rules are:

\[
\begin{array}{ll}
\text{mutual kNN:} & i\sim j \quad\text{if }i\in N_k(j)\text{ and }j\in N_k(i),\\
\text{symmetric kNN:} & i\sim j \quad\text{if }i\in N_k(j)\text{ or }j\in N_k(i),\\
\text{fixed radius:} & i\sim j \quad\text{if }\lVert x_i-x_j\rVert_2\le r,\\
\text{adaptive radius:} & i\sim j \quad\text{if }\lVert x_i-x_j\rVert_2
  \le \delta f(\sigma_i,\sigma_j).
\end{array}
\]

The adaptive rules use `min`, `max`, or geometric-mean choices for $f$; the
geometric-mean case is CkNN. With $\delta=1$, the maximum-radius rule is
equivalent to symmetric kNN under exact, tie-free neighbor calculations. The
package tests this identity across its exact and current zero-epsilon
ANN-backed implementations. Other adaptive rules are alternatives rather than
uniform improvements: CkNN is deliberately sensitive to local sampling
density, and changing $f$ or $\delta$ changes the candidate edge set.
The package also provides intersection-kNN graphs, which retain information
about shared neighborhoods, and iterated graph-geodesic constructions that
replace the ambient distance by shortest paths on a preceding graph.

Every current constructor returns aligned `adj_list` and `weight_list` fields.
Weights are Euclidean edge lengths unless a documented function states
otherwise. Constructors that support pruning and repair also retain lifecycle
branches.

```{r lifecycle-table, echo=FALSE, tab.alt="Six-row table mapping dgraphs lifecycle field names to native, pruned, repaired, and final graph stages. The fields preserve alternative construction histories instead of storing only the selected final graph."}
lifecycle.fields <- data.frame(
  Stage = c(
    "raw_*", "pruned_*", "raw_repaired_*",
    "pruned_repaired_*", "repaired_pruned_*",
    "adj_list, weight_list"
  ),
  Meaning = c(
    "Native construction before pruning or repair.",
    "Optional geometric pruning applied to the native graph.",
    "Native graph after component repair.",
    "Prune-first graph after component repair.",
    "Repair-first graph after pruning.",
    "Final graph selected by the constructor call."
  )
)
accessible.kable(
  lifecycle.fields,
  caption = paste(
    "Lifecycle fields retained by constructors that support pruning and",
    "connectivity repair. Rows distinguish native, operation-order, and final",
    "representations so repair edges are not mistaken for neighborhood edges."
  ),
  row.names = FALSE,
  position = "H"
)
```

This redundancy is deliberate. A minimum-spanning-tree bridge makes a shortest
path finite but does not make that bridge part of the native neighborhood
model. Keeping both stages permits downstream code to distinguish a support
edge from a repair edge.

# A construction-and-assessment workflow

We use a noisy circular point cloud with unequal sample counts on the two
semicircles. Its known circular ordering supplies $D_X$ through shorter-arc
distances, while the density imbalance challenges globally scaled
constructions. A compact constructor comparison uses a nominal neighborhood
scale of seven. Fixed-radius construction uses the median seventh-neighbor
distance.

```{r candidate-graphs}
pairwise.distance <- as.matrix(stats::dist(X))
diag(pairwise.distance) <- Inf
fixed.radius <- stats::median(
  apply(pairwise.distance, 1L, sort, partial = 7L)[7L, ]
)

graphs <- list(
  mutual = create.mknn.graph(
    X, k = 7, connect.components = TRUE
  ),
  symmetric = create.sknn.graph(
    X, k = 7, neighbor.method = "ann", connect.components = TRUE
  ),
  continuous = create.cknn.graph(
    X, k.scale = 7, delta = 1, connect.components = TRUE
  ),
  fixed.radius = create.rknn.graph(
    X, type = "fixed", radius = fixed.radius, connect.components = TRUE
  )
)
```

```{r candidate-summary, echo=FALSE, tab.alt="Four-row table comparing mutual kNN, symmetric kNN, continuous kNN, and fixed-radius graphs at nominal scale seven. It reports native edges and components, bridge edges, and final edges and components. Symmetric kNN is natively connected; each other graph requires one bridge."}
candidate.summary <- do.call(rbind, lapply(names(graphs), function(name) {
  graph <- graphs[[name]]
  data.frame(
    graph = name,
    native_edges = sum(lengths(graph$raw_adj_list)) / 2,
    native_components = graph$n_components_before,
    bridges = graph$n_mst_edges_added,
    final_edges = graph$n_edges,
    final_components = graph$n_components_after
  )
}))
names(candidate.summary) <- c(
  "Graph", "Native edges", "Native comp.", "Bridges",
  "Final edges", "Final comp."
)
accessible.kable(
  candidate.summary,
  caption = paste(
    "Candidate graphs at nominal scale seven. Native edge and component counts",
    "describe the constructor before repair; bridge and final columns show the",
    "effect of component-MST completion. Only symmetric kNN is natively",
    "connected in this example."
  ),
  row.names = FALSE,
  position = "H"
)
```

The edge count is descriptive rather than an objective to maximize. The stage
columns separate the neighborhood rule from component repair. In particular,
the final graph for a disconnected native construction represents a
constructor-plus-repair procedure.

We next vary $k$ for symmetric kNN. The controlled reference permits a stated
selection rule: restrict attention to natively connected candidates and select
the smallest target-normalized graph-geodesic relative RMSE. This use of known
arc distances is specific to the example; an applied analysis needs a
scientifically justified reference or sensitivity criterion.

```{r parameter-sequence, results="hide"}
angle.difference <- abs(outer(theta, theta, "-"))
reference.distance <- pmin(angle.difference, 2 * pi - angle.difference)
k.values <- 3:9

symmetric.sequence <- stats::setNames(lapply(k.values, function(k) {
  create.sknn.graph(
    X,
    k = k,
    neighbor.method = "ann",
    connect.components = TRUE
  )
}), k.values)

symmetric.statistics <- do.call(
  "rbind",
  lapply(symmetric.sequence, function(graph) {
    final.distance <- graph.geodesic.distances(graph)
    native.rel.rms <- if (graph$n_components_before == 1L) {
      isometry.rel.rms.error(
        graph.geodesic.distances(graph, stage = "raw"),
        reference.distance,
        scale = TRUE
      )
    } else {
      NA_real_
    }
    data.frame(
      k = graph$k,
      native_edges = sum(lengths(graph$raw_adj_list)) / 2,
      native_components = graph$n_components_before,
      bridges = graph$n_mst_edges_added,
      native_rel_rmse = native.rel.rms,
      final_rel_rmse = isometry.rel.rms.error(
        final.distance,
        reference.distance,
        scale = TRUE
      )
    )
  })
)

eligible <- symmetric.statistics$native_components == 1L
selected.row <- which(eligible)[which.min(
  symmetric.statistics$final_rel_rmse[eligible]
)]
selected.k <- symmetric.statistics$k[[selected.row]]
selected <- symmetric.sequence[[as.character(selected.k)]]
```

```{r parameter-table, echo=FALSE, tab.alt="Seven-row symmetric-kNN parameter table for k values three through nine. It reports native edges, native components, bridge edges, and target-normalized relative RMSE before and after repair. Native loss is unavailable for disconnected graphs. The first natively connected graph, k seven, has the lowest eligible final loss."}
names(symmetric.statistics) <- c(
  "k", "Native edges", "Native components", "Bridges",
  "Native RelRMSE", "Final RelRMSE"
)
accessible.kable(
  symmetric.statistics,
  caption = paste(
    "Symmetric-kNN candidates over seven neighborhood scales. Relative RMSE",
    "compares graph geodesics with known shorter-arc distances after scale",
    "calibration. Native values are undefined when the native graph is",
    "disconnected. Under the stated rule, k =", selected.k, "is selected."
  ),
  row.names = FALSE,
  digits = 3,
  position = "H"
)
```

```{r workflow-figure, echo=FALSE, fig.pos="!htbp", fig.cap="Selected symmetric-kNN graph on the variable-density circular point cloud. Gray segments are native neighborhood edges and blue points are observations. The selected k = 7 graph is natively connected, so component repair adds no bridge. Selection used the lowest target-normalized graph-geodesic relative RMSE among natively connected candidates.", fig.alt="A circular point cloud with denser sampling on the upper semicircle. Gray symmetric-nearest-neighbor edges connect blue observations around the circle without a component-repair bridge."}
native.edge.matrix <- convert.adjacency.to.edge.matrix(
  selected$raw_adj_list
)$edge.matrix

graphics::plot(
  X,
  asp = 1,
  pch = 19,
  col = "#1F5A94",
  xlab = "Coordinate 1",
  ylab = "Coordinate 2"
)
graphics::segments(
  X[native.edge.matrix[, 1L], 1L], X[native.edge.matrix[, 1L], 2L],
  X[native.edge.matrix[, 2L], 1L], X[native.edge.matrix[, 2L], 2L],
  col = grDevices::adjustcolor("grey30", alpha.f = 0.42)
)
graphics::points(X, pch = 19, col = "#1F5A94")
```

Several graph families also provide plural constructors that return aligned
objects and structural statistics over a parameter sequence. For
intersection-kNN sequences, `compute.stability.metrics()` compares consecutive
graphs using edit distance and Jensen--Shannon divergence between graph-summary
distributions. `build.iknn.graphs.and.selectk()` adds rules based on structural
stability, label mixing, or both, returning the full curves, chosen indices,
and selection evidence.

\FloatBarrier

# Geodesics, summaries, and interoperability

The selected symmetric-kNN graph can be converted directly to an `igraph`
object. Conversion preserves isolated vertices, vertex names when available,
edge lengths in the `weight` attribute, and graph-specific attributes such as
intersection size when present.

```{r conversion}
selected.igraph <- as_igraph(selected)
c(
  vertices = igraph::vcount(selected.igraph),
  edges = igraph::ecount(selected.igraph)
)

degree.summary <- compute.graph.summary.pmf(
  selected,
  summary = "degree_distribution"
)
degree.summary$pmf
```

`graph.geodesic.distances()` computes weighted shortest-path distances. For
the circular example, the shorter angular arc is a known reference. The
isometry summary reports the profiled scale, target-normalized relative RMSE,
distance correlations, multiplicative distortion, shortcut frequency, and
bias over distance ranges.

```{r geodesic-diagnostics, results="hide"}
graph.distance <- graph.geodesic.distances(selected)
geodesic.summary <- summarize.isometry.deviation(
  graph.distance,
  reference.distance,
  scale = TRUE
)
geodesic.diagnostics <- round(unlist(geodesic.summary[c(
  "scale", "rel_rms_error", "rel_abs_error_median", "rel_abs_error_q95",
  "distortion_q05", "distortion_median", "distortion_q95",
  "pearson_cor", "spearman_cor", "shortcut_fraction",
  "short_band_bias", "mid_band_bias", "long_band_bias"
)]), 3)
```

```{r geodesic-table, echo=FALSE, tab.alt="Two-column table of scale-adjusted geodesic diagnostics for the selected symmetric-kNN graph. Rows report calibration scale, target-normalized relative RMSE, relative absolute error, distortion quantiles, distance correlations, shortcut frequency, and distance-range bias."}
accessible.kable(
  data.frame(
    Metric = names(geodesic.diagnostics),
    Value = as.numeric(geodesic.diagnostics)
  ),
  caption = paste(
    "Geodesic fidelity of the selected symmetric-kNN graph relative to known",
    "shorter-arc distances. Relative RMSE is target-normalized after profiled",
    "scale calibration; the remaining rows describe relative error, distortion,",
    "correlation, shortcuts, and distance-range bias."
  ),
  row.names = FALSE,
  position = "H"
)
```

The same representation supports path extraction, endpoint scoring, local
extrema, graph spectra, and spectral embeddings. Their interpretation still
depends on connectivity, repair edges, and geodesic distortion.

# Comparing graph constructions with common diagnostics

## Symmetric-kNN edge parity and timing scope

We compared complete symmetric-kNN graph pipelines using `k = 10` on Gaussian
data with 500, 1,500, or 3,000 observations and 2, 10, or 50 dimensions. The
\CRANpkg{dgraphs} used its bundled ANN kd-tree and returned a lifecycle graph
object. The \CRANpkg{FNN} and \CRANpkg{dbscan} pipelines performed kd-tree
queries and then used the same R helper to union directed neighbor pairs and
compute Euclidean edge weights. Each scenario was warmed up and timed three
times. All `r nrow(pipeline.raw)` retained runs reproduced exactly the same
undirected edge set; ties were absent in these continuous random data.
Figure 2 reads the recorded runs from `data/pipeline-benchmark.csv`; the
publication supplement contains that file, the generating script
`scripts/generate-results.R`, and the software environment record
`data/benchmark-session.txt`.

```{r pipeline-timing, echo=FALSE, fig.pos="!htbp", fig.cap="End-to-end symmetric-kNN graph construction times; points are medians of three runs. FNN was fastest in all tested scenarios. dgraphs remained within the same order of magnitude while also assembling lifecycle diagnostics. Times were measured with one process on the release-R system recorded in the accompanying benchmark metadata.", fig.alt="Faceted line plots of construction time by sample size for dgraphs, FNN, and dbscan in 2, 10, and 50 dimensions. FNN is fastest throughout, while dgraphs and dbscan have similar orders of growth. Runtime increases with sample size and dimension."}
ggplot(
  pipeline.summary,
  aes(x = n, y = median_seconds, color = backend)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~ dimension, labeller = label_both) +
  scale_y_log10() +
  scale_color_manual(values = c(
    dgraphs = "#1F5A94", FNN = "#C44E52", dbscan = "#55A868"
  )) +
  labs(
    x = "Observations",
    y = "Median elapsed seconds (log scale)",
    color = "Pipeline"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

The timing comparison is intentionally scoped to the returned graph, not only
nearest-neighbor lookup. It should not be read as a general benchmark of ANN
libraries. The edge-parity result shows that users can move
between a neighbor-search workflow and the richer \CRANpkg{dgraphs} graph object
without changing the symmetric-kNN rule on these tie-free data. In the largest
scenario ($n=3{,}000$, $p=50$), median elapsed times were
`r sprintf("%.3f", subset(pipeline.summary, backend == "FNN" & n == 3000 & dimension == 50)$median_seconds)`,
`r sprintf("%.3f", subset(pipeline.summary, backend == "dgraphs" & n == 3000 & dimension == 50)$median_seconds)`,
and `r sprintf("%.3f", subset(pipeline.summary, backend == "dbscan" & n == 3000 & dimension == 50)$median_seconds)`
seconds for FNN, \CRANpkg{dgraphs}, and dbscan, respectively.

## Construction families under density imbalance

A second experiment used noisy samples from a unit circle, where the true
shorter-arc distance is known. We varied sample size (100 or 200), angular
density (uniform or uneven), and coordinate noise (0.005 or 0.05), with three
replicates per setting. Mutual-kNN, symmetric-kNN, CkNN, and fixed-radius graphs
used a nominal scale of six neighbors. The fixed radius was the median sixth-
neighbor distance. Each constructor first produced its native graph; exact
component-MST repair was then applied only when that graph was disconnected.

Native component counts and bridge counts evaluate the construction rule
before and during repair. The fidelity plot instead evaluates the connected
final graph. Its target-normalized graph-geodesic relative RMSE therefore
describes the complete constructor-plus-repair procedure whenever bridges were
needed. A native all-pairs geodesic loss is intentionally not reported for a
disconnected graph because its infinite between-component distances do not
define the finite loss used here.

The experiment evaluates sensitivity rather than establishing a general
ordering. Symmetric-kNN attained the lowest scenario median most often and was
natively connected more often than the other families in these settings. The
magnitude of the differences and the ordering among the remaining families
changed with density, noise, and sample size. The raw replicate-level results,
generation script, edge-parity checks, and session metadata accompany the
article so that the comparison can be audited or extended.
Figure 3 and Table 6 read `data/graph-family-benchmark.csv`; the publication
supplement contains that file together with the same generation script and
environment record described above.

```{r family-fidelity, echo=FALSE, fig.pos="!htbp", fig.cap="Final target-normalized graph-geodesic relative RMSE across graph families; lower values indicate closer agreement with known circular arc distances after profiled scale calibration. Faint points are individual replicates and lines connect three-replicate scenario medians. Component-MST repair was applied when a native graph was disconnected, so these values assess the complete constructor-plus-repair procedure. Symmetric kNN attains the lowest scenario median most often in this controlled experiment.", fig.alt="Four faceted plots compare final target-normalized graph-geodesic relative RMSE for mutual kNN, symmetric kNN, continuous kNN, and fixed-radius graphs at 100 and 200 observations. Faint symbols show replicates and connected symbols show medians. Symmetric kNN most often has the lowest median, but the ordering varies by density, noise, and sample size."}
ggplot(
  family.raw,
  aes(
    x = n,
    y = final_rel_rms_error,
    color = family,
    shape = family
  )
) +
  geom_point(
    alpha = 0.28,
    size = 1.5,
    position = position_jitter(width = 2.5, height = 0)
  ) +
  geom_line(
    data = family.summary,
    aes(linetype = family, group = family),
    linewidth = 0.7
  ) +
  geom_point(data = family.summary, size = 2) +
  facet_grid(noise ~ density, labeller = label_both) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Observations",
    y = "Final graph-geodesic relative RMSE",
    color = "Graph family",
    shape = "Graph family",
    linetype = "Graph family"
  ) +
  theme_bw(base_size = 9.5) +
  theme(legend.position = "bottom")
```

```{r family-lifecycle-table, echo=FALSE, tab.alt="Four-row table separating native connectivity and repair from final geodesic fidelity for mutual kNN, symmetric kNN, continuous kNN, and fixed-radius graphs. It reports the number of natively connected replicate runs, median native components, median bridge edges, median final relative RMSE, and the number of scenarios in which each family had the lowest median loss."}
family.lifecycle.display <- family.lifecycle
names(family.lifecycle.display)[
  names(family.lifecycle.display) == "Final RelRMSE"
] <- if (knitr::is_latex_output()) {
  "\\shortstack{Final\\\\RelRMSE}"
} else {
  "Final<br>RelRMSE"
}
names(family.lifecycle.display)[
  names(family.lifecycle.display) == "Lowest-loss scenarios"
] <- if (knitr::is_latex_output()) {
  "\\shortstack{Lowest-loss\\\\scenarios}"
} else {
  "Lowest-loss<br>scenarios"
}
accessible.kable(
  family.lifecycle.display,
  caption = paste(
    "Construction and repair results across 24 replicate runs per graph",
    "family. Connected is the number of natively connected runs; Components",
    "and Bridges are medians. Final relative RMSE includes component-MST repair",
    "when needed. Lowest-loss scenarios counts the lowest median among eight",
    "combinations of sample size, density, and noise; ties, if any, count for",
    "each family."
  ),
  row.names = FALSE,
  digits = 3,
  escape = FALSE,
  position = "H"
)
```

\FloatBarrier

# Scope and limitations

\CRANpkg{dgraphs} assumes finite numerical observations and currently centers
its fast native constructors on Euclidean distance. In the reported tie-free
tests, the current zero-epsilon ANN implementation matched exact/reference
edge sets, including adaptive-radius boundary cases. This is an empirical
parity result, not a general exactness guarantee; deterministic ties can be
backend-dependent and should be tested with `neighbor.method = "exact"` when
tie semantics matter.
Some pruning methods are experimental and are labeled as such in the help
pages. Connectivity repair guarantees a connected final graph when possible,
but it does not validate that the added bridges represent scientific
continuity. A fidelity score for a repaired graph assesses both its native
constructor and the repair rule. Likewise, a stable parameter curve does not
prove that a graph recovers a latent manifold.

The package is therefore most useful when graph construction is treated as a
modeling decision. Users should compare lifecycle stages, inspect repair
edges, examine parameter sequences, and evaluate downstream quantities against
domain-specific references or sensitivity analyses.

The package stops at the first transformation in the opening diagram:
constructing $G$ and comparing $D_X$ with $D_G$. The companion \CRANpkg{grip} work
begins with an already specified graph and assesses coordinate representations
through $D_G$ versus $D_{G,Z}$. Keeping these stages separate makes graph-model
error distinguishable from layout error.

\FloatBarrier

# Summary

\CRANpkg{dgraphs} integrates graph construction, lifecycle-aware repair,
parameter selection, diagnostics, and `igraph` interoperability in one R
package. The software is available from CRAN and developed at
<https://github.com/pgajer/dgraphs>. The end-to-end package vignette can be
opened with `vignette("data-derived-graph-workflow", package = "dgraphs")`.

The article source, benchmark scripts, raw CSV results, and citation
verification record are maintained with the package repository. Workflow,
candidate, parameter-sequence, and geodesic results are evaluated during
rendering; Figures 2--3 and Table 6 summarize the precomputed benchmark CSVs
identified in their respective sections. A companion article on
\CRANpkg{grip} applies the second-stage notation in the opening diagram. The
present article covers construction and $D_X$ versus $D_G$ assessment, whereas
that article begins from a specified graph and concerns $D_G$ versus
$D_{G,Z}$.

# Acknowledgments

Figures and tables were generated with \CRANpkg{ggplot2} [@ggplot2Book] and
\CRANpkg{knitr} [@knitrPackage]. The article was rendered with
\CRANpkg{rmarkdown} [@rmarkdownPackage] and checked with
\CRANpkg{rjtools} [@rjtoolsPackage].
