---
title: "dgraphs: Constructing, Selecting, and Diagnosing Data-Derived Graphs in R"
date: "2026-08-21"
abstract: >
  Data-derived graphs turn numerical observations into vertices and local
  proximity relations into edges. Small choices in the construction rule can
  alter connectivity, graph geodesics, and downstream spectral analyses.
  The dgraphs package provides mutual and symmetric k-nearest-neighbor graphs,
  shared- and intersection-neighbor graphs, fixed and adaptive radius graphs,
  continuous-kNN graphs, graph-geodesic iterations, and minimum-spanning-tree
  completion in one R interface. Constructors preserve native, pruned, and
  connectivity-repaired lifecycle stages instead of silently replacing the
  original graph. Parameter sequences, structural stability summaries,
  geodesic diagnostics, path utilities, spectral embeddings, and conversion
  to igraph support an inspectable workflow from construction to analysis. We
  describe the object design, demonstrate the workflow on a variable-density
  circular point cloud, verify symmetric-kNN edge parity with FNN and dbscan,
  and compare four construction families in a controlled geodesic experiment.
  The package does not select a universally correct graph: its purpose is to
  make graph assumptions and their consequences explicit and reproducible.
draft: true
author:
  - name: Pawel Gajer
    affiliation: University of Maryland School of Medicine
    address:
      - Center for Advanced Microbiome Research and Innovation
      - Institute for Genome Sciences
      - Department of Microbiology and Immunology
      - Baltimore, Maryland, USA
    email: pgajer@gmail.com
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{float}
  \usepackage{placeins}
packages:
  cran:
    - dgraphs
    - FNN
    - dbscan
    - igraph
    - ggplot2
  bioc:
    - bluster
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.7,
  fig.align = "center",
  out.width = "100%",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png",
  dpi = 150
)

library(dgraphs)
library(ggplot2)

stopifnot(packageVersion("dgraphs") >= "0.2.0")

accessible.kable <- function(data, caption, ...) {
  alt <- knitr::opts_current$get("tab.alt")
  if (!is.character(alt) || length(alt) != 1L || !nzchar(trimws(alt))) {
    stop("Accessible tables require a nonempty tab.alt chunk option.")
  }
  if (knitr::is_latex_output()) {
    return(knitr::kable(data, format = "latex", caption = caption, ...))
  }
  escaped.alt <- gsub("&", "&amp;", alt, fixed = TRUE)
  escaped.alt <- gsub("\"", "&quot;", escaped.alt, fixed = TRUE)
  escaped.alt <- gsub("<", "&lt;", escaped.alt, fixed = TRUE)
  escaped.alt <- gsub(">", "&gt;", escaped.alt, fixed = TRUE)
  knitr::kable(
    data,
    format = "html",
    caption = caption,
    table.attr = paste0('aria-label="', escaped.alt, '"'),
    ...
  )
}

pipeline.raw <- utils::read.csv("data/pipeline-benchmark.csv")
family.raw <- utils::read.csv("data/graph-family-benchmark.csv")

pipeline.summary <- stats::aggregate(
  elapsed_seconds ~ backend + n + dimension,
  pipeline.raw,
  stats::median
)
names(pipeline.summary)[names(pipeline.summary) == "elapsed_seconds"] <-
  "median_seconds"

family.summary <- stats::aggregate(
  cbind(
    edges,
    components_before_repair,
    bridges_added,
    normalized_mae,
    shortcut_fraction,
    distance_correlation
  ) ~ family + n + density + noise,
  family.raw,
  stats::median
)

set.seed(20260820)
n.example <- 80L
theta <- sort(c(
  stats::runif(55L, 0, pi),
  stats::runif(25L, pi, 2 * pi)
))
X <- cbind(x = cos(theta), y = sin(theta)) +
  matrix(stats::rnorm(2L * n.example, sd = 0.015), ncol = 2L)
```

# Introduction

A data-derived graph has one vertex for each observation and edges determined
from distances or neighbor relations. Such graphs are inputs to manifold
learning, spectral methods, clustering, trajectory reconstruction, and local
diagnostics. The construction affects the analysis: a fixed-radius graph can
fragment in sparse regions, a symmetric k-nearest-neighbor (kNN) graph can add
links across density transitions, and a mutual-kNN graph can be conservative
enough to disconnect. Consequently, choosing a constructor and choosing its
scale are parts of the statistical analysis.

Several R packages efficiently find nearest neighbors. \CRANpkg{FNN} exposes
kd-tree, cover-tree, and brute-force searches [@fnnPackage], while
\CRANpkg{dbscan} supplies fast spatial indexing together with density-based
clustering methods [@hahsler2019dbscan]. \CRANpkg{igraph} provides a broad
graph-analysis and visualization system [@igraph2023]. These packages are
complementary to \CRANpkg{dgraphs}: the latter focuses on constructing and
comparing data-derived graph families, retaining construction stages, and
diagnosing the geometry of the resulting graph [@dgraphsPackage].

The implemented families have distinct precedents. Shared-nearest-neighbor
ideas trace to @jarvisPatrick1973, mutual-kNN connectivity has been studied for
clustering and outlier detection [@brito1997], and continuous kNN (CkNN) uses
the geometric mean of local scales to adapt to sampling density
[@berrySauer2019]. Optional component repair uses minimum-spanning-tree edges,
linking the graph construction to the classical relationship between minimum
spanning trees and single-linkage structure [@gowerRoss1969]. This article is
about the software design and workflow; it does not propose a new graph rule.

# Positioning and design choices

`dgraphs` is intended for analysts and method developers whose downstream
result depends on the graph construction itself. Its role begins after a data
matrix has been assembled and ends with an explicit graph, its construction
history, and diagnostics that can be passed to another method. Users who need
only nearest-neighbor indices can use \CRANpkg{FNN}; users focused on
density-based clustering can work directly with the kNN, fixed-radius, and
shared-nearest-neighbor objects in \CRANpkg{dbscan} [@hahsler2019dbscan]. The
Bioconductor package \BIOpkg{bluster} constructs kNN and shared-neighbor
`igraph` objects for community detection, allows the neighbor-search backend to
be selected through BiocNeighbors, and supports threaded construction
[@blusterPackage]. These are closer alternatives than a general graph library.
`dgraphs` differs by treating comparisons among graph rules, construction
stages, repair order, and geodesic consequences as part of the returned
analysis rather than as preprocessing for one clustering algorithm.

The public interface follows a `create.<family>.graph()` convention, with
plural constructors for parameter sequences. Constructor results are S3 lists
rather than a new external graph class. Aligned adjacency and weight lists keep
isolated vertices and intermediate branches directly inspectable, serialize
with base R, and make it possible to compare native and repaired graphs without
reconstructing them. The tradeoff is that retaining several lifecycle branches
uses more memory than returning one final edge set. `as_igraph()` provides the
boundary to the broader graph ecosystem instead of reproducing analysis
algorithms already available in \CRANpkg{igraph}.

R performs argument validation, result assembly, and lifecycle bookkeeping;
nearest-neighbor search and the main construction kernels use C or C++ through
Rcpp. The runtime dependencies are limited to FNN, igraph, parallel, and Rcpp,
with RcppEigen used at compilation. The primary constructors are currently
single-process, and the benchmarks in this article use one process. A few
stand-alone utilities expose worker counts, but graph construction itself does
not. Thus, `bluster` is a better fit when threaded kNN or shared-neighbor graph
construction is the main requirement, while `dgraphs` favors deterministic
cross-family diagnostics and lifecycle retention.

For $n$ observations in $p$ dimensions and $m$ retained edges, the input matrix
requires $O(np)$ storage and one adjacency/weight branch requires $O(n+m)$.
Parameter sequences and lifecycle retention increase memory in proportion to
the total number of stored branches and edges. ANN-backed Euclidean searches
avoid an explicit all-pairs distance matrix; exact reference paths and exact
component bridging can require quadratic pair comparisons. Diagnostics that
return all pairwise graph distances necessarily allocate an $n$-by-$n$ result.
For larger data, users can select ANN-backed constructors, request minimal graph
detail when supported, and avoid retaining parameter sequences or all-pairs
diagnostics that are not needed.

# Graph rules and object lifecycle

Let $x_1,\ldots,x_n \in \mathbb{R}^p$, let $N_k(i)$ be the set of $k$
nearest non-self neighbors of vertex $i$, and let $\sigma_i$ be the distance
from $x_i$ to its $k$th neighbor. The principal undirected rules are:

\[
\begin{array}{ll}
\text{mutual kNN:} & i\sim j \quad\text{if }i\in N_k(j)\text{ and }j\in N_k(i),\\
\text{symmetric kNN:} & i\sim j \quad\text{if }i\in N_k(j)\text{ or }j\in N_k(i),\\
\text{fixed radius:} & i\sim j \quad\text{if }\lVert x_i-x_j\rVert_2\le r,\\
\text{adaptive radius:} & i\sim j \quad\text{if }\lVert x_i-x_j\rVert_2
  \le \delta f(\sigma_i,\sigma_j).
\end{array}
\]

The adaptive rules use `min`, `max`, or geometric-mean choices for $f$; the
geometric-mean case is CkNN. The package also provides intersection-kNN graphs,
which retain information about shared neighborhoods, and iterated
graph-geodesic constructions that replace the ambient distance by shortest
paths on a preceding graph.

Every current constructor returns aligned `adj_list` and `weight_list` fields.
Weights are Euclidean edge lengths unless a documented function states
otherwise. Constructors that support pruning and repair also retain lifecycle
branches.

```{r lifecycle-table, echo=FALSE, tab.alt="Six-row table mapping dgraphs lifecycle field names to native, pruned, repaired, and final graph stages. The fields preserve alternative construction histories instead of storing only the selected final graph."}
lifecycle.fields <- data.frame(
  Stage = c(
    "raw_*", "pruned_*", "raw_repaired_*",
    "pruned_repaired_*", "repaired_pruned_*",
    "adj_list, weight_list"
  ),
  Meaning = c(
    "Native construction before pruning or repair.",
    "Optional geometric pruning applied to the native graph.",
    "Native graph after component repair.",
    "Prune-first graph after component repair.",
    "Repair-first graph after pruning.",
    "Final graph selected by the constructor call."
  )
)
accessible.kable(
  lifecycle.fields,
  caption = paste(
    "Lifecycle fields retained by constructors that support pruning and",
    "connectivity repair. Rows distinguish native, operation-order, and final",
    "representations so repair edges are not mistaken for neighborhood edges."
  ),
  row.names = FALSE
)
```

This redundancy is deliberate. A minimum-spanning-tree bridge makes a shortest
path finite but does not make that bridge part of the native neighborhood
model. Keeping both stages permits downstream code to distinguish a support
edge from a repair edge.

# An inspectable construction workflow

We use a noisy circular point cloud with unequal sample counts on the two
semicircles. The known circular ordering supplies a reference geodesic, while
the density imbalance stresses globally scaled constructions.

```{r candidate-graphs}
graphs <- list(
  mutual = create.mknn.graph(
    X, k = 6, connect.components = TRUE
  ),
  symmetric = create.sknn.graph(
    X, k = 6, neighbor.method = "ann", connect.components = TRUE
  ),
  continuous = create.cknn.graph(
    X, k.scale = 6, delta = 1, connect.components = TRUE
  ),
  adaptive.max = create.rknn.graph(
    X, type = "adaptive.radius", k.scale = 6,
    radius.rule = "max", connect.components = TRUE
  )
)
```

```{r candidate-summary, echo=FALSE, tab.alt="Four-row table comparing mutual kNN, symmetric kNN, continuous kNN, and adaptive maximum-radius graphs by edge count, connected components before and after repair, and the number of bridge edges. Symmetric kNN is connected without repair in this example; the other rows show any required bridges."}
candidate.summary <- do.call(rbind, lapply(names(graphs), function(name) {
  graph <- graphs[[name]]
  data.frame(
    graph = name,
    edges = graph$n_edges,
    raw_components = graph$n_components_before,
    final_components = graph$n_components_after,
    bridges = graph$n_mst_edges_added
  )
}))
names(candidate.summary) <- c(
  "Graph", "Edges", "Raw components", "Final components", "Bridges"
)
accessible.kable(
  candidate.summary,
  caption = paste(
    "Candidate graphs for the variable-density circle. Edge counts describe",
    "graph size; component and bridge columns show how much connectivity repair",
    "contributes to each final graph."
  ),
  row.names = FALSE
)
```

The edge count is descriptive rather than an objective to maximize. The
component and bridge columns reveal how much of the final connected graph came
from repair.

```{r workflow-figure, echo=FALSE, fig.pos="H", fig.cap="Continuous-kNN graph on a variable-density circular point cloud. Gray segments are final graph edges; blue points are observations. The constructor retains native and repaired stages separately.", fig.alt="A circular point cloud with denser sampling on the upper semicircle. Gray edges connect nearby blue points around the circle."}
selected <- graphs$continuous
edge.matrix <- convert.adjacency.to.edge.matrix(selected$adj_list)$edge.matrix

graphics::plot(
  X,
  asp = 1,
  pch = 19,
  col = "#1F5A94",
  xlab = "Coordinate 1",
  ylab = "Coordinate 2"
)
graphics::segments(
  X[edge.matrix[, 1L], 1L], X[edge.matrix[, 1L], 2L],
  X[edge.matrix[, 2L], 1L], X[edge.matrix[, 2L], 2L],
  col = grDevices::adjustcolor("grey30", alpha.f = 0.42)
)
graphics::points(X, pch = 19, col = "#1F5A94")
```

Plural constructors make parameter sweeps explicit and return a statistics
table rather than requiring users to reconstruct diagnostics from each object.

```{r parameter-sequence, results="hide"}
radius.sequence <- create.rknn.graphs(
  X,
  k.values = 3:7,
  radius.search = "ann",
  connect.components = TRUE
)
radius.statistics <- radius.sequence$k_statistics[, c(
  "k", "n_edges_before_pruning", "n_components_before",
  "n_mst_edges_added", "n_components_after"
)]
```

```{r parameter-table, echo=FALSE, tab.alt="Five-row parameter-sequence table for k values three through seven, reporting native edges, native connected components, bridge edges, and final connected components. Increasing k adds edges and generally reduces the repair needed for connectivity."}
colnames(radius.statistics) <- c(
  "k", "Native edges", "Native components", "Bridges", "Final components"
)
accessible.kable(
  radius.statistics,
  caption = paste(
    "Adaptive-radius sequence over five neighborhood scales. Native edge and",
    "component counts are reported separately from MST bridges and final",
    "connectivity, exposing the effect of scale before repair."
  ),
  row.names = FALSE
)
```

For intersection-kNN sequences, `compute.stability.metrics()` compares
consecutive graphs using edit distance and Jensen--Shannon divergence between
graph-summary distributions. `build.iknn.graphs.and.selectk()` adds explicit
selection rules based on structural stability, label mixing, or both. These
functions return the full curves, chosen indices, and the evidence used for
selection.

# Geodesics, summaries, and interoperability

The final continuous-kNN graph can be converted directly to an `igraph`
object. Conversion preserves isolated vertices, vertex names when available,
edge lengths in the `weight` attribute, and graph-specific attributes such as
intersection size when present.

```{r conversion}
selected.igraph <- as_igraph(selected)
c(
  vertices = igraph::vcount(selected.igraph),
  edges = igraph::ecount(selected.igraph)
)

degree.summary <- compute.graph.summary.pmf(
  selected,
  summary = "degree_distribution"
)
degree.summary$pmf
```

`graph.geodesic.distances()` computes weighted shortest-path distances. For
the circular example, the shorter angular arc is a known reference. The
isometry diagnostics quantify scale-adjusted stress, distance correlation,
shortcut frequency, and bias over distance ranges.

```{r geodesic-diagnostics, results="hide"}
graph.distance <- graph.geodesic.distances(selected)
angle.difference <- abs(outer(theta, theta, "-"))
reference.distance <- pmin(angle.difference, 2 * pi - angle.difference)

geodesic.diagnostics <- round(isometry.geodesic.diagnostics(
  graph.distance,
  reference.distance
), 3)
```

```{r geodesic-table, echo=FALSE, tab.alt="Two-column table of scale-adjusted geodesic diagnostics for the selected continuous-kNN graph. Rows report calibration scale, relative errors, distance correlations, distortion quantiles, shortcut frequency, and distance-range bias summaries."}
accessible.kable(
  data.frame(
    Metric = names(geodesic.diagnostics),
    Value = as.numeric(geodesic.diagnostics)
  ),
  caption = paste(
    "Geodesic fidelity of the selected continuous-kNN graph relative to known",
    "shorter-arc distances. The rows summarize calibration, error, correlation,",
    "distortion, shortcuts, and distance-range bias."
  ),
  row.names = FALSE
)
```

The same representation supports path extraction, endpoint scoring, local
extrema, graph spectra, and spectral embeddings. Their interpretation still
depends on connectivity, repair edges, and geodesic distortion.

# Comparative validation

## Symmetric-kNN edge parity and timing scope

We compared complete symmetric-kNN graph pipelines using `k = 10` on Gaussian
data with 500, 1,500, or 3,000 observations and 2, 10, or 50 dimensions. The
`dgraphs` pipeline used its bundled ANN kd-tree and returned a lifecycle graph
object. The \CRANpkg{FNN} and \CRANpkg{dbscan} pipelines performed kd-tree
queries and then used the same R helper to union directed neighbor pairs and
compute Euclidean edge weights. Each scenario was warmed up and timed three
times. All `r nrow(pipeline.raw)` retained runs reproduced exactly the same
undirected edge set; ties were absent in these continuous random data.

```{r pipeline-timing, echo=FALSE, fig.pos="H", fig.cap="End-to-end symmetric-kNN graph construction times; points are medians of three runs. FNN was fastest in all tested scenarios. dgraphs remained within the same order of magnitude while also assembling lifecycle diagnostics, and it was slightly faster than the dbscan pipeline at 1,500 and 3,000 observations in 50 dimensions. Times were measured on the recorded single-process R-devel system.", fig.alt="Faceted line plots of construction time by sample size for dgraphs, FNN, and dbscan in 2, 10, and 50 dimensions. FNN is fastest throughout. Runtime increases with sample size and dimension; dgraphs and dbscan are close in 50 dimensions."}
ggplot(
  pipeline.summary,
  aes(x = n, y = median_seconds, color = backend)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  facet_wrap(~ dimension, labeller = label_both) +
  scale_y_log10() +
  scale_color_manual(values = c(
    dgraphs = "#1F5A94", FNN = "#C44E52", dbscan = "#55A868"
  )) +
  labs(
    x = "Observations",
    y = "Median elapsed seconds (log scale)",
    color = "Pipeline"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

The timing comparison is intentionally scoped to the returned graph, not only
nearest-neighbor lookup. It should not be read as a general benchmark of ANN
libraries. The edge-parity result shows that users can move
between a neighbor-search workflow and the richer `dgraphs` graph object
without changing the symmetric-kNN rule on these tie-free data. In the largest
scenario ($n=3{,}000$, $p=50$), median elapsed times were
`r sprintf("%.3f", subset(pipeline.summary, backend == "FNN" & n == 3000 & dimension == 50)$median_seconds)`,
`r sprintf("%.3f", subset(pipeline.summary, backend == "dgraphs" & n == 3000 & dimension == 50)$median_seconds)`,
and `r sprintf("%.3f", subset(pipeline.summary, backend == "dbscan" & n == 3000 & dimension == 50)$median_seconds)`
seconds for FNN, `dgraphs`, and dbscan, respectively.

## Construction families under density imbalance

A second experiment used noisy samples from a unit circle, where the true
shorter-arc distance is known. We varied sample size (100 or 200), angular
density (uniform or uneven), and coordinate noise (0.005 or 0.05), with three
replicates per setting. Mutual-kNN, symmetric-kNN, CkNN, and fixed-radius graphs
used a nominal scale of six neighbors. The fixed radius was the median sixth-
neighbor distance. Exact component-MST repair was enabled for every family.

The experiment evaluates sensitivity rather than establishing a general
ordering. Symmetric-kNN produced the fewest disconnected native graphs and the
smallest displayed median geodesic error in these settings, while the other
families more often required repair. The magnitude of the differences and the
ordering among the remaining families changed with density, noise, and sample
size. The raw replicate-level results, generation script, edge-parity checks,
and session metadata accompany the article so that the comparison can be
audited or extended.

```{r family-fidelity, echo=FALSE, fig.pos="!t", fig.cap="Median normalized absolute geodesic error across graph families; lower values indicate closer agreement with known circular arc distances. Symmetric-kNN has the lowest displayed median in every setting, with the largest separation under uneven sampling. Each point summarizes three replicates, and connectivity repair was enabled for all graphs; the result is specific to this controlled circle experiment.", fig.alt="Four faceted line plots comparing normalized geodesic error for mutual kNN, symmetric kNN, continuous kNN, and fixed-radius graphs at 100 and 200 points. Symmetric kNN is lowest in each density and noise panel, especially under uneven density."}
ggplot(
  family.summary,
  aes(x = n, y = normalized_mae, color = family)
) +
  geom_line(linewidth = 0.65) +
  geom_point(size = 1.8) +
  facet_grid(noise ~ density, labeller = label_both) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    x = "Observations",
    y = "Median normalized absolute error",
    color = "Graph family"
  ) +
  theme_bw(base_size = 9.5) +
  theme(legend.position = "bottom")
```

\FloatBarrier

# Scope and limitations

`dgraphs` assumes finite numerical observations and currently centers its fast
native constructors on Euclidean distance. ANN-backed searches are exact at
the supported tolerance; deterministic ties can still be backend-dependent and
should be tested with `neighbor.method = "exact"` when tie semantics matter.
Some pruning methods are experimental and are labeled as such in the help
pages. Connectivity repair guarantees a connected final graph when possible,
but it does not validate that the added bridges represent scientific
continuity. Likewise, a stable parameter curve does not prove that a graph
recovers a latent manifold.

The package is therefore most useful when graph construction is treated as a
modeling decision. Users should compare lifecycle stages, inspect repair
edges, examine parameter sequences, and evaluate downstream quantities against
domain-specific references or sensitivity analyses.

# Summary and availability

\CRANpkg{dgraphs} integrates graph construction, lifecycle-aware repair,
parameter selection, diagnostics, and `igraph` interoperability in one R
package. The software is available from CRAN and developed at
<https://github.com/pgajer/dgraphs>. The end-to-end package vignette can be
opened with:

```{r vignette-command, eval=FALSE}
vignette("data-derived-graph-workflow", package = "dgraphs")
```

The article source, benchmark scripts, raw CSV results, and citation
verification record are maintained with the package repository.
